---
title: "Untitled"
output: html_document
---

## multiple testing correction
$P(X <= -2 \; | \; \mu=0,\sigma=2)$


We should think of hypothesis testing as a non-errorfree method of making
decisions. There will be times when we declare something significant and accept 

$H_1$ but we will be wrong. 
These decisions are also called "false positives" or "false discoveries", this
is also known as "type I error". Similary, we can fail to reject a hypothesis 
when we actually should. These cases are known as "false negatives", also known
as "type II error". 

The ratio of true negatives to the sum of 
true negatives and false positives ($\frac{TN}{FP+TN}$) is known as specificity.
And we usually want to decrease the FP and get higher specificity. 
The ratio of true positives to the sum of 
true positives and false negatives ($\frac{TP}{TP+FN}$) is known as sensitivity.
And, again we usually want to decrease the FN and get higher sensitiviy. 
Sensitivity is also known as "power of a test" in the context of hypothesis 
testing. More powerful tests will be higly sensitive and will do less type
II errors. For the t-test the power is positively associated with sample size
and the effect size. Higher the sample size, smaller the standard error and 
looking for the larger effect sizes will similary increase the power.

The general summary of these the different combination of the decisions are
included in the table below. 

-------------------------------------------------------------
                 $H_0$ is            $H_1$ is   
                 TRUE,                 TRUE,
                [Gene is NOT          [Gene is
                 differentially      differentially 
                expressed]           expressed]
--------------- -------------------- -------------------- -------------------------
  Accept $H_0$  True Negatives (TN)  False Negatives (FN)  $m_0$: number of truly
  (claim that                          ,type II error       null hypotheses
the gene is not 
differentially
expressed)                                   

  reject $H_0$  False Positives (FP) True Positives (TP)  $m-m_0$: number of 
  (claim that      ,type I error                           truly alternative
the gene is                                                hypotheses
differentially
expressed)
-------------------------------------------------------------


We expect to make more type I errors as the number of tests increase, that 
means we will reject the null hypothesis by mistake. For example, if we 
performe a test the 5% significance level, there is a 5% chance of 
incorrectly rejecting the null hypothesis if the null hypothesis is true. 
However, if we make 1000 tests where all null hypotheses are true for 
each of them, the average number of incorrect rejections is 50. And if we 
apply the rules of probability, there are is almost a 100% chance that
we will have at least one incorrect rejection.
There are multiple statistical techniques to prevent this from happening. 
These techniques generally shrink the P-values obtained from multiple
tests to higher values, if the individual P-value is low enough it survives
this process. The most simple method is just to multiply the individual,
P-value ($p_i$) with the number of tests ($m$): $m \cdot p_i$, this is 
called "bonferonnni correction". However, this is too harsh if you have thousands
of tests. Other methods are developed to remedy this. Those methods 
rely on ranking the P-values and dividing  $m \cdot p_i$ by the 
rank,$i$, :$\frac{m \cdot p_i }{i}$, this is derived from  Benjamini–Hochberg 
procedure. This procedure is developed to control for "False Discovery Rate (FDR)"
, which is proportion of false positives among all significant tests. And in
practical terms, we get the "FDR adjusted P-value" from the procedure described
above. This gives us an estimate of proportion of false discoveries for a given
test.

One final method that is also popular is called the "q-value" 
method and related . This procedure relies on estimating the proportion of true null 
hypotheses from the distribution of raw p-values and using that quantity
to come up with what is called a "q-value". That can be practically defined
as "the proportion of significant features that turn out to be false
leads." A q-value 0.01 would mean 1% of the tests called significant at this 
level will be truly null on average. This is a similar quantity to the one
estimated by "FDR adjusted P-value" method above. Within the genomics community
q-value and FDR adjusted P-value are synoymous. 

In R, the base function `p.adjust()` implements most of the p-value correction 
methods described above. For the q-value, we can use the `qvalue` package from 
Bioconductor. Below we are demonstrating how touse them on a set of simulated 
p-values.The plot shows that Bonferroni correction does a terrible job. FDR(BH) and q-value
approach are better but q-value approach is more permissive than FDR(BH).

```{r}
library(qvalue)
data(hedenfalk)

qvalues <- qvalue(hedenfalk$p)$q
bonf.pval=p.adjust(hedenfalk$p,method ="bonferroni")
fdr.adj.pval=p.adjust(hedenfalk$p,method ="fdr")

plot(hedenfalk$p,qvalues,pch=19,ylim=c(0,1),
     xlab="raw P-values",ylab="adjusted P-values")
points(hedenfalk$p,bonf.pval,pch=19,col="red")
points(hedenfalk$p,fdr.adj.pval,pch=19,col="blue")
legend("bottomright",legend=c("q-value","FDR (BH)","Bonferroni"),
       fill=c("black","blue","red"))
```


generate exp matrix for exercises
```{r}
sd <- 0.2*sqrt(4/rchisq(1000,df=4))
data <- matrix(rnorm(1000*6,sd=sd),1000,6)
rownames(data) <- paste("Gene",1:1000)
data[1:100,4:6] <- data[1:100,4:6] + 2
```


process Xianjun data
```{r}
df=read.table("~/Downloads/TSS-based.cm1w.all.bestbin.lm.0.10.tab.txt",
           header=T,sep="\t")
df2=df[,c("H3k4me3","H3k27me3","measured_log2")]
df3=df2[df$predicted_C==0,]


 cor(df2$H3k4me3,df2$measured_log2)
 cor(df2$H3k27me3,df2$measured_log2)
  
 cor(df3$H3k4me3,df3$measured_log2)
 cor(df3$H3k27me3,df3$measured_log2) 
 
 mod.all=lm(df2$measured_log2 ~ df2$H3k27me3+ df2$H3k4me3 )
 mod.k4=lm(df2$measured_log2 ~  df2$H3k4me3 )
 anova(mod.k4,mod.all,test="F")
 summary(mod.all)
 plot(mod.all,which=1) 
```



```{r}

data=readRDS("StatisticsForGenomics/geneExpMat.rds")

# set groups
group1=1:3
group2=4:6
n1=3
n2=3
dx=rowMeans(data[,group1])-rowMeans(data[,group2])
  
require(matrixStats)

# get the esimate of pooled variance 
stderr <- sqrt( (rowVars(data[,group1])*(n1-1) + rowVars(data[,group2])*(n2-1)) / (n1+n2-2) * ( 1/n1 + 1/n2 ))

# do the shrinking towards median
mod.stderr <- (stderr + median(stderr)) / 2 # moderation in variation

# esimate t statistic with moderated variance
t.mod = dx / mod.stderr

# calculate P-value of rejecting null 
p.mod = 2*pt( -abs(t.mod), n1+n2-2 )

# esimate t statistic without moderated variance
t = dx / stderr

# calculate P-value of rejecting null 
p = 2*pt( -abs(t), n1+n2-2 )

par(mfrow=c(1,2))
hist(p,col="cornflowerblue",border="white",main="",xlab="P-values t-test")
mtext(paste("signifcant tests:",sum(p<0.05))  )
hist(p.mod,col="cornflowerblue",border="white",main="",xlab="P-values mod. t-test")
mtext(paste("signifcant tests:",sum(p.mod<0.05))  )

hist(
p.adjust(p.mod,method ="fdr")
)

hist(
p.adjust(p,method ="fdr")
)

```


```{r}
set.seed(32)

# get 50 X values between 1 and 100
x = runif(50,1,100)

# set b0,b1 and varience (sigma)
b0 = 10
b1 = 2
sigma = 20
# simulate error terms from normal distribution
eps = rnorm(50,0,sigma)
# get y values from the linear equation and addition of error terms
y = b0 + b1*x+ eps

mod=lm(y~x)
sm.mod=summary(mod)

sm.mod$coefficients
sm.mod$coefficients[,4]
cor(x,y)
plot(mod,which=1)
```


```{r}
df=readRDS("StatisticsForGenomics/HistoneModeVSgeneExp.rds")

 cor(df$H3k4me3,df$measured_log2)
 cor(df$H3k27me3,df$measured_log2)
  
 cor(df$H3k4me3,df$measured_log2)
 cor(df$H3k27me3,df$measured_log2) 
 
 mod.all=lm(df$measured_log2 ~ df$H3k27me3+ df$H3k4me3 )
 mod.k4=lm(df$measured_log2 ~  df$H3k4me3 )
 anova(mod.k4,mod.all,test="F")
 summary(mod.all)
 plot(mod.all,which=1) 
```