# Statistics for Genomics {#stats}

```{r setup_statGenomics, include=FALSE}
knitr::opts_chunk$set(echo      = TRUE,
                      message   = FALSE, 
                      error     = FALSE,
                      cache     = TRUE,
                      out.width = "55%",
                      fig.width = 5,
                      fig.align = 'center')
```

This chapter will summarize statistics methods frequently used
in computational genomics. As these fields are continuously evolving,  the
techniques introduced here do not form an exhaustive list but mostly cornerstone methods
that are often and still being used. In addition, we focused on giving intuitive and 
practical understanding of the methods with relevant examples from the field. If you want to dig deeper into statistics and math, beyond what is described
here, we included appropriate references with annotation after each major 
section.

## How to summarize collection of data points: The idea behind statistical distributions
In biology and many other fields, data is collected via experimentation. 
The nature of the experiments and natural variation in biology makes 
it impossible to get the same exact measurements every time you measure something. 
For example, if you are measuring gene expression values for 
a certain gene, say PAX6, and let's assume you are measuring expression
per sample and cell with any method (microarrays, rt-qPCR, etc.). You will not \index{gene expression}
get the same expression value even if your samples are homogeneous, due
to technical bias in experiments or natural variation in the samples. Instead, 
we would like to describe this collection of data some other way
that represents the general properties of the data. Figure \@ref(fig:pax6ReplicatesChp3) shows a sample of
20 expression values from the PAX6 gene.

```{r pax6ReplicatesChp3,fig.align='center',  out.width='50%',echo=FALSE,warning=FALSE,fig.height=5.6,fig.cap="Expression of the PAX6 gene in 20 replicate experiments."}
set.seed(1)
old.par <- par()
a=rnorm(20,mean=6,sd=0.7)
layout(matrix(c(1,2)))
par(fig=c(0,1,0.15,1))
dotchart(a,labels=paste("experiment",20:1),xlim=c(0,12),
         main="PAX6 expression",pch=19)
abline(v=6.13,col="red")

par(fig=c(0,1,0,0.2),mar=c(2,7.65,0.1,2), new=TRUE)

hist(a,xlim=c(0,12),labels = F,main="",  col="red",border="white")

par(old.par)

```

### Describing the central tendency: Mean and median 
As seen in Figure \@ref(fig:pax6ReplicatesChp3), the points from this sample are distributed around 
a central value and the histogram below the dot plot shows the number of points in
each bin. Another observation is that there are some bins that have more points than others. If we want to summarize what we observe, we can try
to represent the collection of data points 
with an expression value that is typical to get, something that represents the
general tendency we observe on the dot plot and the histogram. This value is 
sometimes called the central 
value or central tendency, and there are different ways to calculate such a value. 
In Figure \@ref(fig:pax6ReplicatesChp3), we see that all the values are spread around 6.13 (red line), 
and that is indeed what we call the mean value of this sample of expression values. 
It can be calculated with the following formula $\overline{X}=\sum_{i=1}^n x_i/n$, 
where $x_i$ is the expression value of an experiment and $n$ is the number of 
expression values obtained from the experiments. In R, the `mean()` function will calculate the \index{mean}
mean of a provided vector of numbers. This is called a "sample mean". In reality, there are many more than 20 possible PAX6 expression values (provided each cell is of the 
identical cell type and is in identical conditions). If we had the time and the funding to sample all cells and measure PAX6 expression we would
get a collection of values that would be called, in statistics, a "population". In 
our case, the population will look like the left hand side of the Figure \@ref(fig:pax6MorereplicatesChp3). What we have done with
our 20 data points is that we took a sample of PAX6 expression values from this
population, and calculated the sample mean.

```{r pax6MorereplicatesChp3,out.width='75%',fig.width=6.5,echo=FALSE,warning=FALSE,fig.cap="Expression of all possible PAX6 gene expression measures on all available biological samples (left). Expression of the PAX6 gene from the statistical sample, a random subset from the population of biological samples (right). "}

df=data.frame(x=rnorm(10000,6,0.7))

old.par <- par()
layout(matrix(1:4,ncol=2),heights = c(1,0.5))
par(mar=c( 0, 4.1, 4.1, 2.1))
plot(df[,1],1:nrow(df),pch=19,cex=0.2,yaxt="n",ylab="",xlim=c(2,10),col="blue",
     xaxt="n",xlab="",main="Population")
par(mar=c( 5.1, 4.1, 0, 2.1))
hist(df[,1],xlim=c(2,10),col="blue",border="white",main="",
     xlab="PAX6 expression values")

par(mar=c( 0, 4.1, 4.1, 2.1))
plot(a,1:length(a),pch=19,cex=0.7,yaxt="n",ylab="",xlim=c(2,10),col="red",
     xaxt="n",xlab="",main="Sample")
par(mar=c( 5.1, 4.1, 0, 2.1))
hist(a,xlim=c(2,10),col="red",border="white",main="",
     xlab="PAX6 expression values")
par(old.par)
```

The mean of the population is calculated the same way but traditionally the
Greek letter $\mu$ is used to denote the population mean. Normally, we would not
have access to the population and we will use the sample mean and other quantities 
derived from the sample to estimate the population properties. This is the basic 
idea behind statistical inference, which we will see in action in later 
sections as well. We
estimate the population parameters from the sample parameters and there is some
uncertainty associated with those estimates. We will be trying to assess those
uncertainties and make decisions in the presence of those uncertainties. \index{mean}

We are not yet done with measuring central tendency.
There are other ways to describe it, such as the median value. The 
mean can be affected by outliers easily\index{outliers}. 
If certain values are very high or low compared to the 
bulk of the sample, this will shift mean toward those outliers. However, the median is not affected by outliers. It is simply the value in a distribution where half 
of the values are above and the other half are below. In R, the `median()` function 
will calculate the mean of a provided vector of numbers.  \index{median}Let's create a set of random numbers and calculate their mean and median using
R. 
```{r runifMeanMedChp3}
#create 10 random numbers from uniform distribution 
x=runif(10)
# calculate mean
mean(x)
# calculate median
median(x)
```


### Describing the spread: Measurements of variation
Another useful way to summarize a collection of data points is to measure
how variable the values are. You can simply describe the range of the values, 
such as the minimum and maximum values. You can easily do that in R with the `range()` 
function. A more common way to calculate variation is by calculating something 
called "standard deviation" or the related quantity called "variance". This is a 
quantity that shows how variable the values are. A value around zero indicates 
there is not much variation in the values of the data points, and a high value 
indicates high variation in the values. The variance is the squared distance of 
data points from the mean. Population variance\index{variance} is again a quantity we usually
do not have access to and is simply calculated as follows $\sigma^2=\sum_{i=1}^n \frac{(x_i-\mu)^2}{n}$, where $\mu$ is the population mean, $x_i$ is the $i$th 
data point in the population and $n$ is the population size. However, when we only have access to a sample, this formulation is biased. That means that it
underestimates the population variance, so we make a small adjustment when we
calculate the sample variance, denoted as $s^2$: 

$$
\begin{aligned}
s^2=\sum_{i=1}^n \frac{(x_i-\overline{X})^2}{n-1} && \text{ where $x_i$ is the ith data point and
$\overline{X}$ is the sample mean.}
\end{aligned}
$$


The sample standard deviation is simply the square root of the sample variance, $s=\sqrt{\sum_{i=1}^n \frac{(x_i-\overline{X})^2}{n-1}}$.
The good thing about standard deviation is that it has the same unit as the mean
so it is more intuitive.  


We can calculate the sample standard deviation and variation with the `sd()` and `var()`
functions in R. These functions take a vector of numeric values as input and 
calculate the desired quantities. Below we use those functions on a randomly
generated vector of numbers.
```{r varSdChp3}
x=rnorm(20,mean=6,sd=0.7)
var(x)
sd(x)
```

One potential problem with the variance is that it could be affected by 
outliers.\index{outliers} The points that are too far away from the mean will have a large 
effect on the variance even though there might be few of them.
A way to measure variance that could be less affected by outliers is 
looking at where the bulk of the distribution is. How do we define where the bulk is? 
One common way is to look at the difference between 75th percentile and 25th 
percentile, this effectively removes a lot of potential outliers which will be\index{outliers}
towards the edges of the range of values. 
This is called the interquartile range\index{interquartile range}, and 
can be easily calculated using R via the `IQR()` function and the quantiles of a vector
are calculated with the `quantile()` function.

Let us plot the boxplot for a random vector and also calculate IQR using R. 
In the boxplot (Figure \@ref(fig:boxplot2Chp3)), 25th and 75th percentiles are the edges of the box, and 
the median is marked with a thick line cutting through the box.
```{r IQRChp3}
x=rnorm(20,mean=6,sd=0.7)
IQR(x)
quantile(x)
```

```{r boxplotChp3,eval=FALSE}
boxplot(x,horizontal = T)
```

```{r boxplot2Chp3,fig.height=5.1,out.width='50%',echo=FALSE,warnings=FALSE,message=FALSE,fig.cap="Boxplot showing the 25th percentile and 75th percentile and median for a set of points sampled from a normal distribution with mean=6 and standard deviation=0.7."}

a=quantile(x)[c(2:4)]
boxplot(x,horizontal = T)
text(a[1],1.25,"25th percentile")
text(a[3],1.25,"75th percentile")
```

#### Frequently used statistical distributions
The distributions have parameters (such as mean and variance) that 
summarize them, but also they are functions that assign each outcome of a \index{normal distribution}
statistical experiment to its probability of occurrence.
One distribution that you 
will frequently encounter is the normal distribution or Gaussian distribution.
The normal distribution has a typical "bell-curve" shape
and is characterized by mean and standard deviation. A set of data points 
that
follow normal distribution will mostly be close to the mean 
but spread around it, controlled by the standard deviation parameter. That 
means that if we sample data points from a normal distribution, we are more
likely to sample data points near the mean and sometimes away from the mean. 
The probability of an event occurring is higher if it is nearby the mean.
The effect
of the parameters for the normal distribution can be observed in the following 
plot.

```{r normDistChp3,echo=FALSE,out.width='50%',fig.width=5.1, fig.cap="Different parameters for normal distribution and effect of those on the shape of the distribution"}
plot(function(x) dnorm(x,0,0.5), -5,5,
     main = "",col="red",lwd=2,ylab="P(x)")
curve(dnorm(x,0,1),add=TRUE,col="blue",lwd=2)
curve(dnorm(x,0,2),add=TRUE,col="green",lwd=2)
curve(dnorm(x,-2,1),add=TRUE,col="yellow",lwd=2)
legend("topright",c(expression(paste(mu,"=0, ",sigma,"=0.5")),
                    expression(paste(mu,"=0, ",sigma,"=1")),
                    expression(paste(mu,"=0, ",sigma,"=2")),
                    expression(paste(mu,"=-2, ",sigma,"=1"))),
       col=c("red","blue","green","yellow"),lwd=3,
       bty="n")

```

The normal distribution is often denoted by $\mathcal{N}(\mu,\,\sigma^2)$. When a random variable $X$ is distributed normally with mean $\mu$ and variance $\sigma^2$, we write:

$$X\ \sim\ \mathcal{N}(\mu,\,\sigma^2)$$

The probability
density function of the normal distribution with mean $\mu$ and standard deviation
$\sigma$ is as follows:

$$P(x)=\frac{1}{\sigma\sqrt{2\pi} } \; e^{ -\frac{(x-\mu)^2}{2\sigma^2} } $$

The probability density function gives the probability of observing a value
on a normal distribution defined by the $\mu$ and
$\sigma$ parameters.

Oftentimes, we do not need the exact probability of a value, but we need the
probability of observing a value larger or smaller than a critical value or reference 
point. For example, we might want to know the probability of $X$ being smaller than or
equal to -2 for a normal distribution with mean $0$ and standard deviation $2$: $P(X <= -2 \; | \;  \mu=0,\sigma=2)$. In this case, what we want is the area under the
curve shaded in dark blue. To be able to do that, we need to integrate the probability
density function but we will usually let software do that. Traditionally,
one calculates a Z-score which is simply $(X-\mu)/\sigma=(-2-0)/2= -1$, and
corresponds to how many standard deviations you are away from the mean. 
This is also called "standardization", the corresponding value is distributed in "standard normal distribution" where $\mathcal{N}(0,\,1)$. After calculating the Z-score,
we can look up the area under the curve for the left and right sides of the Z-score in a table, but again, we use software for that.
The tables are outdated when you can use a computer.

Below in Figure \@ref(fig:zscore),  we show the Z-score and the associated probabilities derived
from the calculation above for $P(X <= -2 \; | \;  \mu=0,\sigma=2)$.

```{r zscore,echo=FALSE,message=FALSE,out.width='50%',fig.width=5.1,fig.cap='Z-score and associated probabilities for Z= -1'}

require(mosaic)
xpnorm(c(-2), mean=0, sd=2,lower.tail = TRUE,invisible=T,verbose=FALSE)

```

In R, the family of `*norm` functions (`rnorm`,`dnorm`,`qnorm` and `pnorm`) can
be used to 
operate with the normal distribution, such as calculating probabilities and 
generating random numbers drawn from a normal distribution. We show some of those capabilities below.

```{r drnormChp3}
# get the value of probability density function when X= -2,
# where mean=0 and sd=2
dnorm(-2, mean=0, sd=2)

# get the probability of P(X =< -2) where mean=0 and sd=2
pnorm(-2, mean=0, sd=2)

# get the probability of P(X > -2) where mean=0 and sd=2
pnorm(-2, mean=0, sd=2,lower.tail = FALSE)

# get 5 random numbers from normal dist with  mean=0 and sd=2
rnorm(5, mean=0 , sd=2)

# get y value corresponding to P(X > y) = 0.15 with  mean=0 and sd=2
qnorm( 0.15, mean=0 , sd=2)

```

There are many other distribution functions in R that can be used the same
way. You have to enter the distribution-specific parameters along
with your critical value, quantiles, or number of random numbers depending
on which function you are using in the family. We will list some of those functions below.

- `dbinom` is for the binomial distribution\index{binomial disdistribution}. This distribution is usually used
to model fractional data and binary data. Examples from genomics include
methylation data.

- `dpois` is used for the Poisson distribution and `dnbinom` is used for
the negative binomial distribution. These distributions are used to model count \index{Poisson distribution}
data such as sequencing read counts.

- `df` (F distribution) and `dchisq` (Chi-Squared distribution) are used \index{F distribution}
in relation to the distribution of variation. The F distribution is used to model  \index{Chi-Squared distribution}
ratios of variation and Chi-Squared distribution is used to model 
distribution of variations. You will frequently encounter these in linear models and generalized linear models.


### Precision of estimates: Confidence intervals
When we take a random sample from a population and compute a statistic, such as
the mean, we are trying to approximate the mean of the population. How well this \index{confidence intervals}
sample statistic estimates the population value will always be a
concern. A confidence interval addresses this concern because it provides a 
range of values which will plausibly contain the population parameter of interest.
Normally, we would not have access to a population. If we did, we would not have to estimate the population parameters and their precision.


When we do not have access
to the population, one way to estimate intervals is to repeatedly take samples from the 
original sample with replacement, that is, we take a data point from the sample
we replace, and we take another data point until we have sample size of the 
original sample. Then, we calculate the parameter of interest, in this case the mean, and 
repeat this process a large number of times, such as 1000. At this point, we would have a distribution of re-sampled
means. We can then calculate the 2.5th and 97.5th percentiles and these will
be our so-called 95% confidence interval. This procedure, resampling with replacement to
estimate the precision of population parameter estimates, is known as the __bootstrap resampling__ or __bootstrapping__.\index{bootstrap resampling}

Let's see how we can do this in practice. We simulate a sample
coming from a normal distribution (but we pretend we don't know the 
population parameters). We will estimate the precision
of the mean of the sample using bootstrapping to build confidence intervals, the resulting plot after this procedure is shown in Figure \@ref(fig:bootstrapChp3).

```{r bootstrapChp3,out.width='55%',fig.width=5.1,fig.cap="Precision estimate of the sample mean using 1000 bootstrap samples. Confidence intervals derived from the bootstrap samples are shown with red lines."}
library(mosaic)
set.seed(21)
sample1= rnorm(50,20,5) # simulate a sample

# do bootstrap resampling, sampling with replacement
boot.means=do(1000) * mean(resample(sample1))

# get percentiles from the bootstrap means
q=quantile(boot.means[,1],p=c(0.025,0.975))

# plot the histogram
hist(boot.means[,1],col="cornflowerblue",border="white",
                    xlab="sample means")
abline(v=c(q[1], q[2] ),col="red")
text(x=q[1],y=200,round(q[1],3),adj=c(1,0))
text(x=q[2],y=200,round(q[2],3),adj=c(0,0))

```

If we had a convenient mathematical method to calculate the confidence interval,
we could also do without resampling methods. It turns out that if we take 
repeated 
samples from a population with sample size $n$, the distribution of means
($\overline{X}$) of those samples 
will be approximately normal with mean $\mu$ and standard deviation
$\sigma/\sqrt{n}$. This is also known as the __Central Limit Theorem(CLT)__ and
is one of the most important theorems in statistics. This also means that
$\frac{\overline{X}-\mu}{\sigma\sqrt{n}}$ has a standard normal 
distribution and we can calculate the Z-score, and then we can get 
the percentiles associated with the Z-score. Below, we are showing the 
Z-score
calculation for the distribution of $\overline{X}$, and then
we are deriving the confidence intervals starting with the fact that
the probability of Z being between $-1.96$ and $1.96$ is $0.95$. We then use algebra
to show that the probability that unknown $\mu$ is captured between
$\overline{X}-1.96\sigma/\sqrt{n}$ and $\overline{X}+1.96\sigma/\sqrt{n}$ is $0.95$, which is commonly known as the 95% confidence interval.

$$\begin{array}{ccc}
Z=\frac{\overline{X}-\mu}{\sigma/\sqrt{n}}\\
P(-1.96 < Z < 1.96)=0.95 \\
P(-1.96 < \frac{\overline{X}-\mu}{\sigma/\sqrt{n}} < 1.96)=0.95\\
P(\mu-1.96\sigma/\sqrt{n} < \overline{X} < \mu+1.96\sigma/\sqrt{n})=0.95\\
P(\overline{X}-1.96\sigma/\sqrt{n} < \mu < \overline{X}+1.96\sigma/\sqrt{n})=0.95\\
confint=[\overline{X}-1.96\sigma/\sqrt{n},\overline{X}+1.96\sigma/\sqrt{n}]
\end{array}$$


A 95% confidence interval for the population mean is the most common interval to use, and would \index{confidence interval}
mean that we would expect 95% of the interval estimates to include the 
population parameter, in this case, the mean. However, we can pick any value 
such as 99% or 90%. We can generalize the confidence interval for 
$100(1-\alpha)$ as follows:

$$\overline{X} \pm Z_{\alpha/2}\sigma/\sqrt{n}$$


In R, we can do this using the `qnorm()` function to get Z-scores associated
with ${\alpha/2}$ and ${1-\alpha/2}$. As you can see, the confidence intervals we calculated using CLT are very
similar to the ones we got from the bootstrap for the same sample. For bootstrap we got $[19.21, 21.989]$ and for the CLT-based estimate we got $[19.23638, 22.00819]$.
```{r qnormchp3}
alpha=0.05
sd=5
n=50
mean(sample1)+qnorm(c(alpha/2,1-alpha/2))*sd/sqrt(n)

```

The good thing about CLT is, as long as the sample size is large, regardless of \index{central limit theorem (CLT)}
the population distribution, the distribution of sample means drawn from
that population will always be normal. In Figure \@ref(fig:sampleMeanschp3), we repeatedly 
draw samples 1000 times with sample size $n=10$,$30$, and $100$ from a bimodal,
exponential and a uniform distribution and we are getting sample mean distributions
following normal distribution.

```{r sampleMeanschp3,echo=FALSE,message=FALSE,warning=FALSE,fig.cap="Sample means are normally distributed regardless of the population distribution they are drawn from."}
set.seed(101)
#require(mosaic)
par(mfcol=c(4,3))
par(mar=c(5.1-2,4.1-1,4.1,2.1-2))
d=c(rnorm(1000,mean=10,sd=8),rnorm(1000,mean=40,sd=8))
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=24.8")),cex=0.6)
mtext("bimodal",cex=0.8,line=1)

bimod10=rowMeans(do(1000)*c(rnorm(5,mean=10,sd=8),rnorm(5,mean=40,sd=8)))
bimod30=rowMeans(do(1000)*c(rnorm(15,mean=10,sd=8),rnorm(15,mean=40,sd=8)))
bimod100=rowMeans(do(1000)*c(rnorm(50,mean=10,sd=8),rnorm(50,mean=40,sd=8)))
hist(bimod10,xlim=c(17,33),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=10",side=2,cex=0.8,line=2)
hist(bimod30,xlim=c(17,33),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=30",side=2,cex=0.8,line=2)
hist(bimod100,xlim=c(17,33),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=100",side=2,cex=0.8,line=2)

d=rexp(1000)
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=1")),cex=0.6)
mtext("exponential",cex=0.8,line=1)
mtext("Distributions of different populations",line=2)

exp10 =rowMeans(do(2000)*rexp(10))
exp30 =rowMeans(do(2000)*rexp(30))
exp100=rowMeans(do(2000)*rexp(100))
hist(exp10,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("Sampling distribution of sample means",line=2)
hist(exp30,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(exp100,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")

d=runif(1000)
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=0.5")),cex=0.6)

mtext("uniform",cex=0.8,line=1)
unif10 =rowMeans(do(1000)*runif(10))
unif30 =rowMeans(do(1000)*runif(30))
unif100=rowMeans(do(1000)*runif(100))
hist(unif10,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(unif30,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(unif100,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
```


However, we should note that how we constructed the confidence interval
using standard normal distribution, $N(0,1)$, only works when we know the \index{normal distribution}
population standard deviation. In reality, we usually have only access
to a sample and have no idea about the population standard deviation. If
this is the case, we should estimate the standard deviation using
the sample standard deviation and use something called the _t distribution_ instead  \index{t distribution}
of the standard normal distribution in our interval calculation.  Our confidence interval becomes 
$\overline{X} \pm t_{\alpha/2}s/\sqrt{n}$, with t distribution
parameter $d.f=n-1$, since now  the following quantity is t distributed $\frac{\overline{X}-\mu}{s/\sqrt{n}}$ instead of standard normal distribution.

The t distribution is similar to the standard normal distribution and has mean $0$ but its spread is larger than the normal distribution
especially when the sample size is small, and has one parameter $v$ for
the degrees of freedom, which is $n-1$ in this case. Degrees of freedom
is simply the number of data points minus the number of parameters estimated.\index{degrees of freedom}Here we are estimating the mean from the data, therefore the degrees of freedom is $n-1$. The resulting distributions are shown in Figure \@ref(fig:tdistChp3).
```{r tdistChp3,echo=FALSE,warning=FALSE,message=FALSE,out.width='60%',fig.cap="Normal distribution and t distribution with different degrees of freedom. With increasing degrees of freedom, the t distribution approximates the normal distribution better."}
plot(function(x) dnorm(x,0,1), -4,4,
     main = "",col="red",lwd=2,ylab="P(x)")
curve(dt(x,1),add=TRUE,col="orange",lwd=2)
curve(dt(x,3),add=TRUE,col="green",lwd=2)
curve(dt(x,10),add=TRUE,col="blue",lwd=2)
legend("topright",c(expression(paste("N(",mu,"=0, ",sigma,"=1)")),
                    expression(paste(v,"=1")),
                    expression(paste(v,"=3")),
                    expression(paste(v,"=10"))),
       col=c("red","orange","green","blue"),lwd=3,
       bty="n")

```

## How to test for differences between samples
Oftentimes we would want to compare sets of samples. Such comparisons include
if wild-type samples have different expression compared to mutants or if healthy
samples are different from disease samples in some measurable feature (blood count,
gene expression, methylation of certain loci). Since there is variability in our
measurements, we need to take that into account when comparing the sets of samples.
We can simply subtract the means of two samples, but given the variability
of sampling, at the very least we need to decide a cutoff value for differences
of means; small differences of means can be explained by random chance due to
sampling. That means we need to compare the difference we get to a value that
is typical to get if the difference between two group means were only due to 
sampling. If you followed the logic above, here we actually introduced two core 
ideas of something called "hypothesis testing", which is simply using 
statistics to \index{hypothesis testing}
determine the probability that a given hypothesis (Ex: if two sample sets
are from the same population or not) is true. Formally, expanded version of those two core ideas are as follows:

1. Decide on a hypothesis to test, often called the "null hypothesis" ($H_0$). In our 
 case, the hypothesis is that there is no difference between sets of samples. An "alternative hypothesis" ($H_1$) is that there is a difference between the
 samples.
2. Decide on a statistic to test the truth of the null hypothesis.
3. Calculate the statistic.
4. Compare it to a reference value to establish significance, the P-value. Based on that, either reject or not reject the null hypothesis, $H_0$.


### Randomization-based testing for difference of the means 
There is one intuitive way to go about this. If we believe there are no 
differences between samples, that means the sample labels (test vs. control or 
healthy vs. disease) have no meaning. So, if we randomly assign labels to the 
samples and calculate the difference of the means, this creates a null 
distribution for $H_0$ where we can compare the real difference and 
measure how unlikely it is to get such a value under the expectation of the
null hypothesis. We can calculate all possible permutations to calculate
the null distribution. However, sometimes that is not very feasible and the
equivalent approach would be generating the null distribution by taking a 
smaller number of random samples with shuffled group membership.

Below, we are doing this process in R. We are first simulating two samples 
from two different distributions.
These would be equivalent to gene expression measurements obtained under 
different conditions. Then, we calculate the differences in the means
and do the randomization procedure to get a null distribution when we 
assume there is no difference between samples, $H_0$. We then calculate how
often we would get the original difference we calculated under the 
assumption that $H_0$ is true. The resulting null distribution and the original value is shown in Figure \@ref(fig:randomTestchp3).


```{r randomTestchp3,out.width='60%',fig.cap="The null distribution for differences of means obtained via randomization. The original difference is marked via the blue line. The red line marks the value that corresponds to P-value of 0.05"}
set.seed(100)
gene1=rnorm(30,mean=4,sd=2)
gene2=rnorm(30,mean=2,sd=2)
org.diff=mean(gene1)-mean(gene2)
gene.df=data.frame(exp=c(gene1,gene2),
                  group=c( rep("test",30),rep("control",30) ) )


exp.null <- do(1000) * diff(mosaic::mean(exp ~ shuffle(group), data=gene.df))
hist(exp.null[,1],xlab="null distribution | no difference in samples",
     main=expression(paste(H[0]," :no difference in means") ),
     xlim=c(-2,2),col="cornflowerblue",border="white")
abline(v=quantile(exp.null[,1],0.95),col="red" )
abline(v=org.diff,col="blue" )
text(x=quantile(exp.null[,1],0.95),y=200,"0.05",adj=c(1,0),col="red")
text(x=org.diff,y=200,"org. diff.",adj=c(1,0),col="blue")
p.val=sum(exp.null[,1]>org.diff)/length(exp.null[,1])
p.val

```

After doing random permutations and getting a null distribution, it is possible to get a confidence interval for the distribution of difference in means.
This is simply the $2.5th$ and $97.5th$ percentiles of the null distribution, and 
directly related to the P-value calculation above.


### Using t-test for difference of the means between two samples
We can also calculate the difference between means using a t-test\index{t-test}. Sometimes we will have too few data points in a sample to do a meaningful
randomization test, also randomization takes more time than doing a t-test.
This is a test that depends on the t distribution\index{t distribution}. The line of thought follows
from the CLT and we can show differences in means are t distributed.
There are a couple of variants of the t-test for this purpose. If we assume
the population variances are equal we can use the following version

$$t = \frac{\bar {X}_1 - \bar{X}_2}{s_{X_1X_2} \cdot \sqrt{\frac{1}{n_1}+\frac{1}{n_2}}}$$
where 
$$s_{X_1X_2} = \sqrt{\frac{(n_1-1)s_{X_1}^2+(n_2-1)s_{X_2}^2}{n_1+n_2-2}}$$
In the first equation above, the quantity is t distributed with $n_1+n_2-2$ degrees of freedom. We can calculate the quantity and then use software 
to look for the percentile of that value in that t distribution, which is our P-value. When we cannot assume equal variances, we use "Welch's t-test"
which is the default t-test in R and also works well when variances and
the sample sizes are the same. For this test we calculate the following 
quantity:

$$t = \frac{\overline{X}_1 - \overline{X}_2}{s_{\overline{X}_1 - \overline{X}_2}}$$
where 
$$s_{\overline{X}_1 - \overline{X}_2} = \sqrt{\frac{s_1^2 }{ n_1} + \frac{s_2^2 }{n_2}}$$

and the degrees of freedom equals to

$$\mathrm{d.f.} = \frac{(s_1^2/n_1 + s_2^2/n_2)^2}{(s_1^2/n_1)^2/(n_1-1) + (s_2^2/n_2)^2/(n_2-1)}
$$

Luckily, R does all those calculations for us. Below we will show the use of `t.test()` function in R. We will use it on the samples we simulated 
above.
```{r welchTtest}
# Welch's t-test
stats::t.test(gene1,gene2)

# t-test with equal variance assumption
stats::t.test(gene1,gene2,var.equal=TRUE)
```

A final word on t-tests: they generally assume a population where samples coming 
from them have a normal
distribution, however it is been shown t-test can tolerate deviations from
normality, especially, when two distributions are moderately skewed in the 
same direction. This is due to the central limit theorem, which says that the means of 
samples will be distributed normally no matter the population distribution 
if sample sizes are large.

### Multiple testing correction

We should think of hypothesis testing as a non-error-free method of making \index{multiple testing correction}
decisions. There will be times when we declare something significant and accept 
$H_1$ but we will be wrong. 
These decisions are also called "false positives" or "false discoveries", and are also known as "type I errors". Similarly, we can fail to reject a hypothesis 
when we actually should. These cases are known as "false negatives", also known
as "type II errors". 

The ratio of true negatives to the sum of 
true negatives and false positives ($\frac{TN}{FP+TN}$) is known as specificity.
And we usually want to decrease the FP and get higher specificity. 
The ratio of true positives to the sum of 
true positives and false negatives ($\frac{TP}{TP+FN}$) is known as sensitivity.
And, again, we usually want to decrease the FN and get higher sensitivity. 
Sensitivity is also known as the "power of a test" in the context of hypothesis 
testing. More powerful tests will be highly sensitive and will have fewer type
II errors. For the t-test, the power is positively associated with sample size
and the effect size. The larger the sample size, the smaller the standard error, and 
looking for the larger effect sizes will similarly increase the power.

The general summary of these different decision combinations are
included in the table below. 

-------------------------------------------------------------
                 $H_0$ is            $H_1$ is   
                 TRUE,                 TRUE,
                [Gene is NOT          [Gene is
                 differentially      differentially 
                expressed]           expressed]
--------------- -------------------- -------------------- -------------------------
  Accept $H_0$  True Negatives (TN)  False Negatives (FN)  $m_0$: number of truly
  (claim that                          ,type II error       null hypotheses
the gene is not 
differentially
expressed)                                   

  reject $H_0$  False Positives (FP) True Positives (TP)  $m-m_0$: number of 
  (claim that      ,type I error                           truly alternative
the gene is                                                hypotheses
differentially
expressed)
-------------------------------------------------------------


We expect to make more type I errors as the number of tests increase, which
means we will reject the null hypothesis by mistake. For example, if we 
perform a test at the 5% significance level, there is a 5% chance of 
incorrectly rejecting the null hypothesis if the null hypothesis is true. 
However, if we make 1000 tests where all null hypotheses are true for 
each of them, the average number of incorrect rejections is 50. And if we 
apply the rules of probability, there is almost a 100% chance that
we will have at least one incorrect rejection.
There are multiple statistical techniques to prevent this from happening. 
These techniques generally push the P-values obtained from multiple
tests to higher values; if the individual P-value is low enough it survives
this process. The simplest method is just to multiply the individual
P-value ($p_i$) by the number of tests ($m$), $m \cdot p_i$. This is 
called "Bonferroni correction". However, this is too harsh if you have thousands
of tests. Other methods are developed to remedy this. Those methods 
rely on ranking the P-values and dividing  $m \cdot p_i$ by the 
rank, $i$, :$\frac{m \cdot p_i }{i}$, which is derived from the Benjamini–Hochberg \index{P-value}
procedure. This procedure is developed to control for "False Discovery Rate (FDR)"
, which is the proportion of false positives among all significant tests. And in
practical terms, we get the "FDR-adjusted P-value" from the procedure described
above. This gives us an estimate of the proportion of false discoveries for a given
test. To elaborate, p-value of 0.05 implies that 5% of all tests will be false positives. An FDR-adjusted p-value of 0.05 implies that 5% of significant tests will be false positives. The FDR-adjusted P-values will result in a lower number of false positives.

One final method that is also popular is called the "q-value" 
method and related to the method above. This procedure relies on estimating the proportion of true null 
hypotheses from the distribution of raw p-values and using that quantity
to come up with what is called a "q-value", which is also an FDR-adjusted P-value [@Storey2003-nv]. That can be practically defined
as "the proportion of significant features that turn out to be false
leads." A q-value 0.01 would mean 1% of the tests called significant at this \index{q-value}
level will be truly null on average. Within the genomics community
q-value and FDR adjusted P-value are synonymous although they can be 
calculated differently.

In R, the base function `p.adjust()` implements most of the p-value correction 
methods described above. For the q-value, we can use the `qvalue` package from 
Bioconductor. Below we demonstrate how to use them on a set of simulated 
p-values. The plot in Figure \@ref(fig:multtest) shows that Bonferroni correction does a terrible job. FDR(BH) and q-value
approach are better but, the q-value approach is more permissive than FDR(BH).

```{r multtest,out.width='60%',fig.cap="Adjusted P-values via different methods and their relationship to raw P-values"}
library(qvalue)
data(hedenfalk)

qvalues <- qvalue(hedenfalk$p)$q
bonf.pval=p.adjust(hedenfalk$p,method ="bonferroni")
fdr.adj.pval=p.adjust(hedenfalk$p,method ="fdr")

plot(hedenfalk$p,qvalues,pch=19,ylim=c(0,1),
     xlab="raw P-values",ylab="adjusted P-values")
points(hedenfalk$p,bonf.pval,pch=19,col="red")
points(hedenfalk$p,fdr.adj.pval,pch=19,col="blue")
legend("bottomright",legend=c("q-value","FDR (BH)","Bonferroni"),
       fill=c("black","blue","red"))
```

### Moderated t-tests: Using information from multiple comparisons
In genomics, we usually do not do one test but many, as described above. That means we\index{moderated t-test}
may be able to use the information from the parameters obtained from all 
comparisons to influence the individual parameters. For example, if you have many variances
calculated for thousands of genes across samples, you can force individual 
variance estimates to shrink toward the mean or the median of the distribution
of variances. This usually creates better performance in individual variance
estimates and therefore better performance in significance testing, which
depends on variance estimates. How much the values are shrunk toward a common
value depends on the exact method used. These tests in general are called moderated
t-tests or shrinkage t-tests. One approach popularized by Limma software is
to use so-called "Empirical Bayes methods"\index{empirical Bayes methods}. The main formulation in these
methods is $\hat{V_g} = aV_0 + bV_g$, where $V_0$ is the background variability and $V_g$ is the individual variability. Then, these methods estimate $a$ and $b$ in various ways to come up with a "shrunk" version of the variability, $\hat{V_g}$. Bayesian inference can make use of prior knowledge to make inference about properties of the data. In a Bayesian viewpoint,
the prior knowledge, in this case variability of other genes, can be used to calculate the variability of an individual gene. In our
case, $V_0$ would be the prior knowledge we have on the variability of 
the genes and we
use that knowledge to influence our estimate for the individual genes.

Below we are simulating a gene expression matrix with 1000 genes, and 3 test 
and 3 control groups. Each row is a gene, and in normal circumstances we would
like to find differentially expressed genes. In this case, we are simulating
them from the same distribution, so in reality we do not expect any differences.
We then use the adjusted standard error estimates in empirical Bayesian spirit but, in a very crude way. We just shrink the gene-wise standard error estimates towards the median with equal $a$ and $b$ weights. That is to say, we add the individual estimate to the 
median of the standard error distribution from all genes and divide that quantity by 2. So if we plug that into the above formula, what we do is:

$$ \hat{V_g} = (V_0 + V_g)/2 $$

In the code below, we are avoiding for loops or apply family functions
by using vectorized operations. The code below samples gene expression values from a hypothetical distribution. Since all the values come from the same distribution, we do not expect differences between groups. We then calculate moderated and unmoderated t-test statistics and plot the P-value distributions for tests. The results are shown in Figure \@ref(fig:modTtestChp3).
```{r modTtestChp3, out.width='60%',fig.width=8,fig.cap="The distributions of P-values obtained by t-tests and moderated t-tests"}
set.seed(100)

#sample data matrix from normal distribution

gset=rnorm(3000,mean=200,sd=70)
data=matrix(gset,ncol=6)

# set groups
group1=1:3
group2=4:6
n1=3
n2=3
dx=rowMeans(data[,group1])-rowMeans(data[,group2])
  
require(matrixStats)

# get the estimate of pooled variance 
stderr = sqrt( (rowVars(data[,group1])*(n1-1) + 
       rowVars(data[,group2])*(n2-1)) / (n1+n2-2) * ( 1/n1 + 1/n2 ))

# do the shrinking towards median
mod.stderr = (stderr + median(stderr)) / 2 # moderation in variation

# estimate t statistic with moderated variance
t.mod <- dx / mod.stderr

# calculate P-value of rejecting null 
p.mod = 2*pt( -abs(t.mod), n1+n2-2 )

# estimate t statistic without moderated variance
t = dx / stderr

# calculate P-value of rejecting null 
p = 2*pt( -abs(t), n1+n2-2 )

par(mfrow=c(1,2))
hist(p,col="cornflowerblue",border="white",main="",xlab="P-values t-test")
mtext(paste("signifcant tests:",sum(p<0.05))  )
hist(p.mod,col="cornflowerblue",border="white",main="",
     xlab="P-values mod. t-test")
mtext(paste("signifcant tests:",sum(p.mod<0.05))  )

```


```{block2, note-text3, type='rmdtip'}

__Want to know more ?__

- Basic statistical concepts  
    - "Cartoon guide to statistics" by Gonick & Smith [@gonick2005cartoon]. Provides central concepts depicted as cartoons in a funny but clear and accurate manner.
    - "OpenIntro Statistics" [@diez2015openintro] (Free e-book http://openintro.org). This book provides fundamental statistical concepts in a clear and easy way. It includes R code.
- Hands-on statistics recipes with R
    - "The R book" [@crawley2012r]. This is the main R book for anyone interested in statistical concepts and their application in R. It requires some background in statistics since the main focus is applications in R.
- Moderated tests
    - Comparison of moderated tests for differential expression [@de2010benchmark] http://bmcbioinformatics.biomedcentral.com/articles/10.1186/1471-2105-11-17
    - Limma method developed for testing differential expression between genes using a moderated test [@smyth2004linear] http://www.statsci.org/smyth/pubs/ebayes.pdf

```



## Relationship between variables: Linear models and correlation
In genomics, we would often need to measure or model the relationship between 
variables. We might want to know about expression of a particular gene in liver 
in relation to the dosage of a drug that patient receives. Or, we may want to know 
DNA methylation of a certain locus in the genome in relation to the age of the sample donor. Or, we might be interested in the relationship between histone
modifications and gene expression\index{histone modification}. Is there a linear relationship, the more \index{gene expression}
histone modification the more the gene is expressed ? 

In these
situations and many more, linear regression or linear models can be used to \index{linear regression}
model the relationship with a "dependent" or "response" variable (expression or 
methylation
in the above examples) and one or more "independent" or "explanatory" variables (age, drug dosage or histone modification in the above examples). Our simple linear model has the 
following components. 

$$  Y= \beta_0+\beta_1X + \epsilon $$ 


 In the equation above, $Y$ is the response variable and $X$ is the explanatory 
 variable. $\epsilon$ is the mean-zero error term. Since the line fit will not 
 be able to precisely predict the $Y$ values, there will be some error associated
 with each prediction when we compare it to the original $Y$ values. This error
 is captured in the $\epsilon$ term. We can alternatively write the model as 
 follows to emphasize that the model approximates $Y$, in this case notice that we removed the $\epsilon$ term: $Y \sim \beta_0+\beta_1X$.
 
 
 The plot below in Figure \@ref(fig:histoneLmChp3) shows the relationship between
 histone modification (trimethylated forms of histone H3 at lysine 4, aka H3K4me3)
 and gene expression for 100 genes. The blue line is our model with estimated
 coefficients ($\hat{y}=\hat{\beta}_0 + \hat{\beta}_1X$, where $\hat{\beta}_0$
 and $\hat{\beta}_1$ are the estimated values of  $\beta_0$ and
 $\beta_1$, and $\hat{y}$ indicates the prediction). The red lines indicate the individual
 errors per data point, indicated as $\epsilon$ in the formula above. 
 
```{r histoneLmChp3,echo=FALSE,warning=FALSE,message=FALSE,error=FALSE,results='hide',out.width='60%',fig.cap="Relationship between histone modification score and gene expression. Increasing histone modification, H3K4me3, seems to be associated with increasing gene expression. Each dot is a gene"}
set.seed(31)
x1 <- runif(100,10,200)


b0 <- 17
b1 <- 0.5

sigma <- 15

eps <- rnorm(100,0,sigma)
y <- b0 + b1*x1 + eps

plot(x1,y,ylim=c(0,160),xlim=c(0,220),pch=20,
     ylab="Gene Expression",xlab="Histone modification score")
mod1=lm(y~x1)
abline(mod1,col="blue")

# calculate residuals and predicted values
res <- signif(residuals(mod1), 5)
pre <- predict(mod1) # plot distances between points and the regression line
segments(x1, y, x1, pre, col="red")



```
 
There could be more than one explanatory variable. We then simply add more $X$ 
and $\beta$ to our model. If there are two explanatory variables our model
will look like this:
 
 $$  Y= \beta_0+\beta_1X_1 +\beta_2X_2 + \epsilon $$ 
 
In this case, we will be fitting a plane rather than a line. However, the fitting
process which we will describe in the later sections will not change for our 
gene expression problem. We can introduce one more histone modification, H3K27me3. We will then have a linear model with 2 explanatory variables and the 
fitted plane will look like the one in Figure \@ref(fig:histoneLm2chp3). The gene expression values are shown
as dots below and above the fitted plane. Linear regression and its extensions which make use of other distributions (generalized linear models) \index{generalized linear model} are central in computational genomics for statistical tests. We will see more of how regression is used in statistical hypothesis testing for computational genomics in Chapters \@ref(rnaseqanalysis) and  \@ref(bsseq).

```{r histoneLm2chp3,echo=FALSE,out.width='65%',warning=FALSE,message=FALSE,fig.cap="Association of gene expression with H3K4me3 and H3K27me3 histone modifications."}
set.seed(32)
x2 <- runif(100,10,200)

b2 <- -0.3
sigma <- 15

eps <- rnorm(100,0,sigma)
y2 <- b0 + b1*x1 + b2*x2+ eps


library(plot3D)


mod1=lm(y2~x1+x2)

# predict on x-y grid, for surface
x1.pred <- seq(min(x1), max(x1), length.out = 30)
x2.pred <- seq(min(x2), max(x2), length.out = 30)
xy <- expand.grid(x1 = x1.pred, 
                  x2= x2.pred)
 
y.pred <- matrix (nrow = 30, ncol = 30, 
  data = predict(mod1, newdata = data.frame(xy), interval = "prediction"))
 
# predicted z-values, fitted points for droplines to surface
fitpoints <- predict(mod1) 
 
scatter3D(z = y2, x = x1, y = x2, pch = 19, cex = 0.4,colvar=sign(residuals(mod1)),
          col = c("magenta","red"), 
      theta = 20, phi = 40, ticktype = "simple", bty= "g",
      xlab = "H3K4me3", ylab = "H3K27me3", 
      zlab ="Gene exp." ,r=sqrt(5),
      surf = list(x = x1.pred, y = x2.pred, z = y.pred, 
                  facets = NA, fit = fitpoints,col="blue"),
      colkey = FALSE)


```
 

 

#### Matrix notation for linear models
We can naturally have more explanatory variables than just two. The formula 
below has $n$ explanatory variables.
 
 $$Y= \beta_0+\beta_1X_1+\beta_2X_2 +  \beta_3X_3 + .. + \beta_nX_n +\epsilon$$

If there are many variables, it would be easier
to write the model in matrix notation. The matrix form of linear model with
two explanatory variables will look like the one 
below. The first matrix would be our data matrix. This contains our explanatory 
variables and a column of 1s. The second term is a column vector of $\beta$ 
values. We also add a vector of error terms, $\epsilon$s, to the matrix multiplication.
 
$$
 \mathbf{Y} = \left[\begin{array}{rrr}
1 & X_{1,1} & X_{1,2} \\
1 & X_{2,1} & X_{2,2} \\
1 & X_{3,1} & X_{3,2} \\
1 & X_{4,1} & X_{4,2}
\end{array}\right]
%
\left[\begin{array}{rrr}
\beta_0 \\
\beta_1 \\
\beta_2 
\end{array}\right]
% 
+
\left[\begin{array}{rrr}
\epsilon_1 \\
\epsilon_2 \\ 
\epsilon_3 \\ 
\epsilon_4
\end{array}\right]
$$
 
The multiplication of the data matrix and $\beta$ vector and addition of the
error terms simply results in the following set of equations per data point:

$$
\begin{aligned}
Y_1= \beta_0+\beta_1X_{1,1}+\beta_2X_{1,2} +\epsilon_1 \\
Y_2= \beta_0+\beta_1X_{2,1}+\beta_2X_{2,2} +\epsilon_2 \\
Y_3= \beta_0+\beta_1X_{3,1}+\beta_2X_{3,2} +\epsilon_3 \\
Y_4= \beta_0+\beta_1X_{4,1}+\beta_2X_{4,2} +\epsilon_4 
\end{aligned}
$$


This expression involving the multiplication of the data matrix, the
$\beta$ vector and vector of error terms ($\epsilon$) 
could be simply written as follows.

$$Y=X\beta + \epsilon$$

In the equation, above $Y$ is the vector of response variables, $X$ is the
data matrix, and $\beta$ is the vector of coefficients.
This notation is more concise and often used in scientific papers. However, this
also means you need some understanding of linear algebra to follow the math 
laid out in such resources. 
 
 
### How to fit a line
At this point a major question is left unanswered: How did we fit this line?
We basically need to define $\beta$ values in a structured way.
There are multiple ways of understanding how
to do this, all of which converge to the same 
end point. We will describe them one by one.

#### The cost or loss function approach
This is the first approach and in my opinion is easiest to understand. \index{cost function}
We try to optimize a function, often called the "cost function" or "loss function". \index{loss function}
The cost function
is the sum of squared differences between the predicted $\hat{Y}$ values from our model
and the original $Y$ values. The optimization procedure tries to find $\beta$ values \index{optimization}
that minimize this difference between the reality and predicted values.
 
 $$min \sum{(y_i-(\beta_0+\beta_1x_i))^2}$$

Note that this is related to the error term, $\epsilon$, we already mentioned
above. We are trying to minimize the squared sum of $\epsilon_i$ for each data
point. We can do this minimization by a bit of calculus. 
The rough algorithm is as follows:

1. Pick a random starting point, random $\beta$ values.
2. Take the partial derivatives of the cost function to see which direction is
  the way to go in the cost function.
3. Take a step toward the direction that minimizes the cost function.
    - Step size is a parameter to choose, there are many variants.
4. Repeat step 2,3 until convergence.

This is the basis of the "gradient descent" algorithm.\index{gradient descent} With the help of partial 
derivatives we define a "gradient" on the cost function and follow that through
multiple iterations until convergence, meaning until the results do not 
improve defined by a margin. The algorithm usually converges to optimum $\beta$
values. In Figure \@ref(fig:3dcostfunc), we show the cost function over various $\beta_0$ and $\beta_1$
values for the histone modification and gene expression data set. The algorithm
will pick a point on this graph and traverse it incrementally based on the 
derivatives and converge to the bottom of the cost function "well". Such optimization methods are the core of machine learning methods we will cover later in Chapters \@ref(unsupervisedLearning) and
\@ref(supervisedLearning).


```{r 3dcostfunc,fig.height=4,echo=FALSE,warning=FALSE,message=FALSE,fig.cap="Cost function landscape for linear regression with changing beta values. The optimization process tries to find the lowest point in this landscape by implementing a strategy for updating beta values toward the lowest point in the landscape."}

require(plot3D)

costfun<-function(b0,b1,x,y){
  0.5*sum((y-b0-b1*x1)^2)
}

b0=seq(15,35,l=200)
b1=seq(0.1,0.8,l=200)
M=mesh(b0,b1)
e=mapply(costfun,b0=M$x,b1=M$y, 
         MoreArgs=list(x=x1,y=y))

e2=mapply(costfun,b0=b0, 
         MoreArgs=list(b1=0.45,x=x1,y=y))     

e3=mapply(costfun,b1=b1, 
         MoreArgs=list(b0=23,x=x1,y=y)) 

#scatter3Drgl(z = e, x = M$x, y = M$y, pch = 19, cex = 0.4,
#      theta = 50, phi = 20, ticktype = "detailed", bty= "g",
#      xlab = "b0", ylab = "b1", zlab ="error" ,
#      colkey = FALSE)

#surf3Drgl(z = matrix(e,ncol=100), x =  M$x , y =  M$y,
#          contour=TRUE)
par(mfrow=c(1,2))
surf3D(z = matrix(e,ncol=200), x =  M$x , y =  M$y,bty="f",
       zlab="cost function",ylab=expression(beta[0]),
       phi=10,d=10,cex.lab = 0.7,padj=1,
       xlab=expression(beta[1]),colkey = FALSE)
image2D(z=matrix(e,ncol=200),contour=T,colkey = FALSE,
        xlab=expression(beta[1]),ylab=expression(beta[0]))
par(mfrow=c(1,1))

```




#### Not cost function but maximum likelihood function
We can also think of this problem from a more statistical point of view. In \index{maximum likelihood estimation}
essence, we are looking for best statistical parameters, in this 
case $\beta$ values, for our model that are most likely to produce such a 
scatter of data points given the explanatory variables. This is called the
"maximum likelihood" approach. The approach assumes that a given response variable $y_i$ follows a normal distribution with mean $\beta_0+\beta_1x_i$ and \index{variance} variance $s^2$. Therefore the probability of observing any given $y_i$ value is dependent on the $\beta_0$ and  $\beta_1$ values. Since $x_i$, the explanatory variable, is fixed within our data set, we can maximize the probability of observing any given $y_i$ by varying  $\beta_0$ and  $\beta_1$ values. The trick is to find $\beta_0$ and  $\beta_1$ values that maximizes the probability of observing all the response variables in the dataset given the explanatory variables. The probability of observing a response variable $y_i$ with assumptions we described above is shown below. Note that this assumes variance is constant and $s^2=\frac{\sum{\epsilon_i}}{n-2}$ is an unbiased estimation for population variance, $\sigma^2$.\index{variance}

 $$P(y_{i})=\frac{1}{s\sqrt{2\pi} }e^{-\frac{1}{2}\left(\frac{y_i-(\beta_0 + \beta_1x_i)}{s}\right)^2}$$
 
Following from the probability equation above, the likelihood function (shown as $L$ below) for 
linear regression is \index{linear regression} multiplication of $P(y_{i})$ for all data points.

$$L=P(y_1)P(y_2)P(y_3)..P(y_n)=\prod\limits_{i=1}^n{P_i}$$

This can be simplified to the following equation by some algebra, assumption of normal distribution, and taking logs (since it is 
easier to add than multiply).

$$ln(L) = -nln(s\sqrt{2\pi}) - \frac{1}{2s^2} \sum\limits_{i=1}^n{(y_i-(\beta_0 + \beta_1x_i))^2} $$

As you can see, the right part of the function is the negative of the cost function
defined above. If we wanted to optimize this function we would need to take the derivative of
the function with respect to the $\beta$ parameters. That means we can ignore the 
first part since there are no $\beta$ terms there. This simply reduces to the 
negative of the cost function. Hence, this approach produces exactly the same 
result as the cost function approach. The difference is that we defined our 
problem
within the domain of statistics. This particular function has still to be optimized. This can be done with some calculus without the need for an 
iterative approach. 

The maximum likelihood approach also opens up other possibilities for regression. For the case above, we assumed that the points around the mean are distributed by normal distribution. However, there are other cases where this assumption may not hold. For example, for the count data the mean and variance relationship is not constant; the higher the mean counts, the higher the variance. In these cases, the regression framework with maximum likelihood estimation can still be used. We simply change the underlying assumptions about the distribution and calculate the likelihood with a new distribution in mind,
and maximize the parameters for that likelihood. This gives way to "generalized linear model"\index{generalized linear model} approach where errors for the response variables can have other distributions than normal distribution. We will see examples of these generalized linear models in Chapter \@ref(rnaseqanalysis) and \@ref(bsseq).




#### Linear algebra and closed-form solution to linear regression
The last approach we will describe is the minimization process using linear \index{linear regression}
algebra. If you find this concept challenging, feel free to skip it, but scientific publications and other books frequently use matrix notation and linear algebra to define and solve regression problems. In this case, we do not use an iterative approach. Instead, we will
minimize the cost function by explicitly taking its derivatives with respect to 
$\beta$'s and setting them to zero. This is doable by employing linear algebra 
and matrix calculus. This approach is also called "ordinary least squares". We \index{ordinary least squares regression}
will not 
show the whole derivation here, but the following expression
is what we are trying to minimize in matrix notation, which is basically a
different notation of the same minimization problem defined above. Remember
$\epsilon_i=Y_i-(\beta_0+\beta_1x_i)$

$$
\begin{aligned}
\sum\epsilon_{i}^2=\epsilon^T\epsilon=(Y-{\beta}{X})^T(Y-{\beta}{X}) \\
=Y^T{Y}-2{\beta}^T{Y}+{\beta}^TX^TX{\beta}
\end{aligned}
$$
After rearranging the terms, we take the derivative of $\epsilon^T\epsilon$ 
with respect to $\beta$, and equalize that to zero. We then arrive at 
the following for estimated $\beta$ values, $\hat{\beta}$:

$$\hat{\beta}=(X^TX)^{-1}X^TY$$


This requires you to calculate the inverse of the $X^TX$ term, which could 
be slow for large matrices. Using an iterative approach over the cost function 
derivatives will be faster for larger problems.
The linear algebra notation is something you will see in the papers 
or other resources often. If you input the data matrix X and solve the $(X^TX)^{-1}$
, 
you get the following values for $\beta_0$ and $\beta_1$ for simple regression \index{linear regression}. However, we should note that this simple linear regression case can easily 
be solved algebraically without the need for matrix operations. This can be done
by taking the derivative of $\sum{(y_i-(\beta_0+\beta_1x_i))^2}$ with respect to
$\beta_1$, rearranging the terms and equalizing the derivative to zero.

$$\hat{\beta_1}=\frac{\sum{(x_i-\overline{X})(y_i-\overline{Y})}}{ \sum{(x_i-\overline{X})^2} }$$ 
$$\hat{\beta_0}=\overline{Y}-\hat{\beta_1}\overline{X}$$


#### Fitting lines in R
After all this theory, you will be surprised how easy it is to fit lines in R.
This is achieved just by the `lm()` function, which stands for linear models. Let's do this
for a simulated data set and plot the fit. The first step is to simulate the 
data. We will decide on $\beta_0$ and $\beta_1$ values. Then we will decide
on the variance parameter, $\sigma$, to be used in simulation of error terms,\index{variance}
$\epsilon$. We will first find $Y$ values, just using the linear equation
$Y=\beta0+\beta_1X$, for 
a set of $X$ values. Then, we will add the error terms to get our simulated values.
```{r getFittinLineData}
# set random number seed, so that the random numbers from the text
# is the same when you run the code.
set.seed(32)

# get 50 X values between 1 and 100
x = runif(50,1,100)

# set b0,b1 and variance (sigma)
b0 = 10
b1 = 2
sigma = 20
# simulate error terms from normal distribution
eps = rnorm(50,0,sigma)
# get y values from the linear equation and addition of error terms
y = b0 + b1*x+ eps


```
Now let us fit a line using the `lm()` function. The function requires a formula, and
optionally a data frame. We need to pass the following expression within the
`lm()` function, `y~x`, where `y` is the simulated $Y$ values and `x` is the explanatory variables $X$. We will then use the `abline()` function to draw the fit. The resulting plot is shown in Figure \@ref(fig:geneExpLinearModel).
```{r geneExpLinearModel,out.width='60%',fig.cap="Gene expression and histone modification score modeled by linear regression."}
mod1=lm(y~x)

# plot the data points
plot(x,y,pch=20,
     ylab="Gene Expression",xlab="Histone modification score")
# plot the linear fit
abline(mod1,col="blue")
```


### How to estimate the error of the coefficients  
Since we are using a sample to estimate the coefficients, they are
not exact; with every random sample they will vary. In Figure \@ref(fig:regCoeffRandomSamples), we
take multiple samples from the population and fit lines to each 
sample; with each sample the lines slightly change. We are overlaying the
points and the lines for each sample on top of the other samples. When we take 200 samples and fit lines for each of them, the line fits are 
variable. And,
we get a normal-like distribution of $\beta$ values with a defined mean \index{linear regression}
and standard deviation, which is called standard error of the
coefficients.

```{r regCoeffRandomSamples,message=FALSE,warning=FALSE,echo=FALSE,fig.cap="Regression coefficients vary with every random sample. The figure illustrates the variability of regression coefficients when regression is done using a sample of data points. Histograms depict this variability for $b_0$ and $b_1$ coefficients."}
set.seed(31)
b0 <- 17
b1 <- 0.5
sigma <- 30
eps <- rnorm(100,0,sigma)
x <- runif(100,10,200)
y <- b0 + b1*x + eps

par(mfrow=c(2,3))
require(scales)
plot(x,y,ylim=c(0,160),xlim=c(0,220),pch=20,
     col=scales::alpha("blue", 0.1),main="1 sample")

mod1=lm(y~x)
abline(mod1,col="red")

simXY<-function(b1,b0,sigma){
  eps <- rnorm(100,0,sigma)
  x1 <- runif(100,10,200)
  y <- b0 + b1*x1 + eps
  list(x1,y)
}

plot(x,y,ylim=c(0,160),xlim=c(0,220),pch=20,col=scales::alpha("blue", 0.1),main="2 samples")
abline(lm(y~x),col="red")
xy=simXY(b1,b0,sigma)
points(xy[[1]],xy[[2]],pch=20,col=scales::alpha("blue", 0.1))
abline(lm(xy[[2]]~xy[[1]]),col="red")

plot(x,y,ylim=c(0,160),xlim=c(0,220),pch=20,col=scales::alpha("blue", 0.1),main="10 samples")
abline(lm(y~x),col="red")
for(i in 1:9){
xy=simXY(b1,b0,sigma)
points(xy[[1]],xy[[2]],pch=20,col=scales::alpha("blue", 0.05))
abline(lm(xy[[2]]~xy[[1]]),col="red")
}

plot(x,y,ylim=c(0,160),xlim=c(0,220),pch=20,col=scales::alpha("blue", 0.1),main="200 samples")
abline(lm(y~x),col="red")
b0s=numeric(200)
b1s=numeric(200)

for(i in 1:200){
  xy=simXY(b1,b0,sigma)
  points(xy[[1]],xy[[2]],pch=20,col=scales::alpha("blue", 0.01))
  modxy=lm(xy[[2]]~xy[[1]])
  abline(modxy,col="red")
  b0s[i]=coef(modxy)[1]
  b1s[i]=coef(modxy)[2]
}
mb0s=round(mean(b0s),2)
sdb0s=round(sd(b0s),2)
hist(b0s,breaks=10,xlab=expression(beta[0]),
     col="cornflowerblue",border="white",
     main=bquote(bar(x) == .(mb0s) ~~ s == .(sdb0s))
     )

mb1s=round(mean(b1s),2)
sdb1s=round(sd(b1s),2)
hist(b1s,breaks=10,xlab=expression(beta[1]),
     col="cornflowerblue",border="white",
     main=bquote(bar(x) == .(mb1s) ~~ s == .(sdb1s))
     )

```

Normally, we will not have access to the population to do repeated sampling, 
model fitting, and estimation of the standard error for the coefficients. But
there is statistical theory that helps us infer the population properties from
the sample. When we assume that error terms have constant variance and mean zero
, we can model the uncertainty in the regression coefficients, $\beta$s. 
The estimates for standard errors of $\beta$s for simple regression are as \index{linear regression}
follows and shown without derivation.

$$
\begin{aligned}
s=RSE=\sqrt{\frac{\sum{(y_i-(\beta_0+\beta_1x_i))^2}}{n-2}  } =\sqrt{\frac{\sum{\epsilon^2}}{n-2}  } \\
SE(\hat{\beta_1})=\frac{s}{\sqrt{\sum{(x_i-\overline{X})^2}}} \\
SE(\hat{\beta_0})=s\sqrt{ \frac{1}{n} + \frac{\overline{X}^2}{\sum{(x_i-\overline{X})^2} }  }
\end{aligned}
$$

Notice that that $SE(\beta_1)$ depends on the estimate of variance of
residuals shown as $s$ or __Residual Standard Error (RSE)__.\index{variance} \index{residuals}
Notice also the standard error depends on the spread of $X$. If $X$ values have more \index{Residual Standard Error (RSE)}
variation, the standard error will be lower. This intuitively makes sense since if the 
spread of $X$ is low, the regression line will be able to wiggle more 
compared to a regression line that is fit to the same number of points but
covers a greater range on the X-axis.


The standard error estimates can also be used to calculate confidence intervals and test
hypotheses, since the following quantity, called t-score, approximately follows a
t-distribution with $n-p$ degrees of freedom, where $n$ is the number
of data points and $p$ is the number of coefficients estimated.

$$ \frac{\hat{\beta_i}-\beta_test}{SE(\hat{\beta_i})}$$

Often, we would like to test the null hypothesis if a coefficient is equal to
zero or not. For simple regression, this could mean if there is a relationship
between the explanatory variable and the response variable. We would calculate the 
t-score as follows $\frac{\hat{\beta_i}-0}{SE(\hat{\beta_i})}$, and compare it
to the t-distribution with $d.f.=n-p$ to get the p-value.


We can also 
calculate the uncertainty of the regression coefficients using confidence 
intervals, the range of values that are likely to contain $\beta_i$. The 95% 
confidence interval for $\hat{\beta_i}$  is 
$\hat{\beta_i}$ ± $t_{0.975}SE(\hat{\beta_i})$.
$t_{0.975}$ is the 97.5% percentile of 
the t-distribution with $d.f. = n – p$.


In R, the `summary()` function will test all the coefficients for the null hypothesis
$\beta_i=0$. The function takes the model output obtained from the `lm()` 
function. To demonstrate this, let us first get some data. The procedure below
simulates data to be used in a regression setting and it is useful to examine \index{linear regression}
what the linear model expects to model the data.
```{r extraLinarModelDataGeneration,echo=FALSE,warning=FALSE,message=FALSE}
# set random number seed, so that the random numbers from the text
# is the same when you run the code.
set.seed(32)

# get 100 X values between 1 and 100
x = runif(100,10,200)
# set b0,b1 and variance (sigma)
b0 = 17
b1 = 0.5
sigma = 30
# simulate error terms from normal distribution
eps = rnorm(100,0,sigma)
# get y values from the linear equation and addition of error terms
y = b0 + b1*x+ eps


```

Since we have the data, we can build our model and call the `summary` function.
We will then use the `confint()` function to get the confidence intervals on the 
coefficients and the `coef()` function to pull out the estimated coefficients from
the model.
```{r confintLM}
mod1=lm(y~x)
summary(mod1)

# get confidence intervals 
confint(mod1)

# pull out coefficients from the model
coef(mod1)
```
The `summary()` function prints out an extensive list of values. 
The "Coefficients" section has the estimates, their standard error, t score,
and the p-value from the hypothesis test $H_0:\beta_i=0$. As you can see, the
estimate we get for the coefficients and their standard errors are close to
the ones we get from repeatedly sampling and getting a distribution of 
coefficients. This is statistical inference at work, so we can estimate the 
population properties within a certain error using just a sample.





### Accuracy of the model
If you have observed the table output of the `summary()` function, you must have noticed there are some other outputs, such as "Residual standard error",
"Multiple R-squared" and "F-statistic". These are metrics that are useful 
for assessing the accuracy of the model. We will explain them one by one. 

__RSE__ is simply the square-root of\index{Residual Standard Error (RSE)}
the sum of squared error terms, divided by degrees of freedom, $n-p$. For the simple 
linear regression case, degrees of freedom is $n-2$. Sum of the squares of the error terms is also
called the __"Residual sum of squares"__, RSS. \index{Residual sum of squares (RSS)}So the RSE is 
calculated as follows:

$$ s=RSE=\sqrt{\frac{\sum{(y_i-\hat{Y_i})^2 }}{n-p}}=\sqrt{\frac{RSS}{n-p}}$$

The RSE is a way of assessing the model fit. The larger the RSE the worse the 
model is. However, this is an absolute measure in the units of $Y$ and we have nothing to 
compare against. One idea is that we divide it by the RSS of a simpler model
for comparative purposes. That simpler model is in this case is the model
with the intercept, $\beta_0$. A very bad model will have close to zero 
coefficients for explanatory variables, and the RSS of that model
will be close to the RSS of the model with only the intercept. In such
a model the intercept will be equal to $\overline{Y}$. As it turns out, the RSS of the model with 
just the intercept is called the _"Total Sum of Squares" or TSS_. A good model will have a low $RSS/TSS$. The metric $R^2$ uses these quantities to calculate a score between 0 and 1, and the closer to 1, the better the model. Here is how 
it is calculated:

$$R^2=1-\frac{RSS}{TSS}=\frac{TSS-RSS}{TSS}=1-\frac{RSS}{TSS}$$

The $TSS-RSS$ part of the formula is often referred to as "explained variability" in 
the model. The bottom part is for "total variability". With this interpretation, the higher 
the "explained variability", the better the model. For simple linear regression
with one explanatory variable, the square root of $R^2$ is a quantity known
as the absolute value of the correlation coefficient, which can be calculated for any pair of variables, not only 
the
response and the explanatory variables. _Correlation_ is the general measure of \index{correlation}
linear
relationship between two variables. One 
of the most popular flavors of correlation is the Pearson correlation coefficient. Formally, it is the 
_covariance_ of X and Y divided by multiplication of standard deviations of \index{covariance}
X and Y. In R, it can be calculated with the `cor()` function.

$$ 
r_{xy}=\frac{cov(X,Y)}{\sigma_x\sigma_y}
      =\frac{\sum\limits_{i=1}^n (x_i-\bar{x})(y_i-\bar{y})}
            {\sqrt{\sum\limits_{i=1}^n (x_i-\bar{x})^2 \sum\limits_{i=1}^n (y_i-\bar{y})^2}}
$$
In the equation above, $cov$ is the covariance; this is again a measure of 
how much two variables change together, like correlation. If two variables \index{covariance}
show similar behavior, they will usually have a positive covariance value. If they have opposite behavior, the covariance will have a negative value.
However, these values are boundless. A normalized way of looking at
covariance is to divide covariance by the multiplication of standard
errors of X and Y. This bounds the values to -1 and 1, and as mentioned
above, is called Pearson correlation coefficient. The values that change in a similar manner will have a positive coefficient, the values that change in \index{correlation}
an opposite manner will have a negative coefficient, and pairs that do not have
a linear relationship will have $0$ or near $0$ correlation. In 
Figure \@ref(fig:CorCovar), we are showing $R^2$, the correlation 
coefficient, and covariance for different scatter plots.

```{r CorCovar,fig.width=17,fig.height=5,out.width = "100%",echo=FALSE,warning=FALSE,message=FALSE, fig.cap="Correlation and covariance for different scatter plots."}
set.seed(31)
x=runif(50,min=5,max=75)
eps=rnorm(50,sd=50)

par(mfrow=c(1,5))
par(mar=c(5.1,1.1,4.1,0.1))
y3=5+5*x
plot(x,y3,xlab="",xaxt="n",yaxt="n",col="cornflowerblue",pch=19,
     cex.main=1.5,
     main=
bquote(R^2 == .(cor(x,y3)^2) ~~ r == .(cor(x,y3)) ~~ Cov== .(cov(x,y3)) ) )

y3=5+5*x+eps
plot(x,y3,xlab="",xaxt="n",yaxt="n",col="cornflowerblue",pch=19,cex.main=1.5,
     main=
bquote(R^2 == .(round(cor(x,y3)^2,2)) ~~ r == .(round(cor(x,y3),2)) ~~ Cov== .(cov(x,y3)) ) )

y3=rep(5,length(x))+eps
plot(x,y3,xlab="",xaxt="n",yaxt="n",col="cornflowerblue",pch=19,cex.main=1.5,
     main=
bquote(R^2 == .(round(cor(x,y3)^2,2)) ~~ r == .(round(cor(x,y3),2)) ~~ Cov== .(cov(x,y3)) ) )

y3=5-5*x+eps
plot(x,y3,xlab="",xaxt="n",yaxt="n",col="cornflowerblue",pch=19,cex.main=1.5,
     main=
bquote(R^2 == .(round(cor(x,y3)^2,2)) ~~ r == .(round(cor(x,y3),2)) ~~ Cov== .(cov(x,y3)) ) )

y3=5-5*x
plot(x,y3,xlab="",xaxt="n",yaxt="n",col="cornflowerblue",pch=19,cex.main=1.5,
     main=
bquote(R^2 == .(round(cor(x,y3)^2,2)) ~~ r == .(round(cor(x,y3),2)) ~~ Cov== .(cov(x,y3)) ) )
```

For simple linear regression, correlation can be used to assess the model. However, this becomes useless as a measure of general accuracy
if there is more than one explanatory variable as in multiple linear regression. In that case, $R^2$ is a measure 
of accuracy for the model. Interestingly, the square of the 
correlation of predicted values
and original response variables ($(cor(Y,\hat{Y}))^2$ ) equals $R^2$ for \index{$R^2$}
multiple linear regression.\index{linear regression}


The last accuracy measure, or the model fit in general we are going to explain is _F-statistic_. This is a quantity that depends on the RSS and TSS again. It can also answer one important question that other metrics cannot easily answer. That question is whether or not any of the explanatory
variables have predictive value or in other words if all the explanatory variables are zero. We can write the null hypothesis as follows:

$$H_0: \beta_1=\beta_2=\beta_3=...=\beta_p=0 $$ 

where the alternative is:

$$H_1: \text{at least one } \beta_i \neq 0 $$ 

Remember that $TSS-RSS$ is analogous to "explained variability" and the RSS is
analogous to "unexplained variability". For the F-statistic, we divide explained variance by
unexplained variance. Explained variance is just the $TSS-RSS$ divided 
by degrees of freedom, and unexplained variance is the RSE.
The ratio will follow the F-distribution
with two parameters, the degrees of freedom for the explained variance and
the degrees of freedom for the unexplained variance. The F-statistic for a linear model is calculated as follows.

$$F=\frac{(TSS-RSS)/(p-1)}{RSS/(n-p)}=\frac{(TSS-RSS)/(p-1)}{RSE} \sim F(p-1,n-p)$$

If the variances are the same, the ratio will be 1, and when $H_0$ is true, then
it can be shown that expected value of $(TSS-RSS)/(p-1)$ will be $\sigma^2$, which is estimated by the RSE. So, if  the variances are significantly different, 
the ratio will need to be significantly bigger than 1.
If the ratio is large enough we can reject the null hypothesis. To assess that,
we need to use software or look up the tables for F statistics with calculated
parameters. In R, function `qf()` can be used to calculate critical value of the
ratio. Benefit of the F-test over
looking at significance of coefficients one by one is that we circumvent
multiple testing problem. If there are lots of explanatory variables
at least 5% of the time (assuming we use 0.05 as P-value significance 
cutoff), p-values from coefficient t-tests will be wrong\index{t-test}. In summary, F-test is a better choice for testing if there is any association
between the explanatory variables and the response variable.




### Regression with categorical variables
An important feature of linear regression is that categorical variables can
be used as explanatory variables, this feature is very useful in genomics
where explanatory variables can often be categorical. To put it in 
context, in our histone modification \index{histone modification} example we can also include if 
promoters have CpG islands or not as a variable. In addition, in 
differential gene expression, we usually test the difference between
different conditions, which can be encoded as categorical variables in
a linear regression.\index{linear regression} We can sure use the t-test for that as well if there are only 2 conditions, but if there are more conditions and other variables
to control for, such as age or sex of the samples, we need to take those 
into account for our statistics, and the t-test alone cannot handle such 
complexity. In addition, when we have categorical variables we can also
have numeric variables in the model and we certainly do not have to include
only one type of variable in a model. 

The simplest model with categorical variables includes two levels that
can be encoded in 0 and 1. Below, we show linear regression with categorical variables. We then plot the fitted line. This plot is shown in Figure \@ref(fig:LMcategorical).
```{r LMcategorical,out.width='50%',fig.cap="Linear model with a categorical variable coded as 0 and 1."}
set.seed(100)
gene1=rnorm(30,mean=4,sd=2)
gene2=rnorm(30,mean=2,sd=2)
gene.df=data.frame(exp=c(gene1,gene2),
                  group=c( rep(1,30),rep(0,30) ) )

mod2=lm(exp~group,data=gene.df)
summary(mod2)
require(mosaic)
plotModel(mod2)
```

We can even compare more levels, and we do not even have to encode them
ourselves. We can pass categorical variables to the `lm()` function.
```{r LMcatcompare}

gene.df=data.frame(exp=c(gene1,gene2,gene2),
                  group=c( rep("A",30),rep("B",30),rep("C",30) ) 
                  )

mod3=lm(exp~group,data=gene.df)
summary(mod3)
```


### Regression pitfalls
In most cases one should look at the error terms (residuals) vs. the fitted
values plot. Any structure in this plot indicates problems such as 
non-linearity, correlation of error terms\index{correlation}, non-constant variance or
unusual values driving the fit. Below we briefly explain the potential
issues with the linear regression.

##### Non-linearity
If the true relationship is far from linearity, prediction accuracy
is reduced and all the other conclusions are questionable. In some cases,
transforming the data with $logX$, $\sqrt{X}$, and $X^2$ could resolve
the issue.

##### Correlation of explanatory variables
If the explanatory variables are correlated that could lead to something  
known as multicollinearity. When this happens SE estimates of the coefficients will be too large. This is usually observed in time-course
data.

##### Correlation of error terms
This assumes that the errors of the response variables are uncorrelated with each other. If they are, the confidence intervals of the coefficients 
might be too narrow.


##### Non-constant variance of error terms 
This means that different response variables have the same variance in their errors, regardless of the values of the predictor variables. If 
the errors are not constant (ex: the errors grow as X values increase), this
will result in unreliable estimates in standard errors as the model
assumes constant variance. Transformation of data, such as
$logX$ and $\sqrt{X}$, could help in some cases.

##### Outliers and high leverage points
Outliers are extreme values for Y and high leverage points are unusual\index{outliers}
X values. Both of these extremes have the power to affect the fitted line
and the standard errors. In some cases (Ex: if there are measurement errors), they can be 
removed from the data for a better fit.

```{block2, statsLinMod, type='rmdtip'}

__Want to know more ?__

- Linear models and derivations of equations including matrix notation
    - _Applied Linear Statistical Models_ by Kutner, Nachtsheim, et al. [@kutner2003applied]
    - _Elements of Statistical Learning_ by Hastie & Tibshirani [@friedman2001elements]
    - _An Introduction to Statistical Learning_ by James, Witten, et al. [@james2013introduction]

```

## Exercises 

### How to summarize collection of data points: The idea behind statistical distributions

1. Calculate the means and variances 
of the rows of the following simulated data set, and plot the distributions
of means and variances using `hist()` and `boxplot()` functions. [Difficulty: **Beginner/Intermediate**]  
```{r getDataChp3Ex,eval=FALSE}
set.seed(100)

#sample data matrix from normal distribution
gset=rnorm(600,mean=200,sd=70)
data=matrix(gset,ncol=6)
```


2. Using the data generated above, calculate the standard deviation of the
distribution of the means using the `sd()` function. Compare that to the expected
standard error obtained from the central limit theorem keeping in mind the
population parameters were  $\sigma=70$ and $n=6$. How does the estimate from the random samples change if we simulate more data with
`data=matrix(rnorm(6000,mean=200,sd=70),ncol=6)`? [Difficulty: **Beginner/Intermediate**] 

3. Simulate 30 random variables using the `rpois()` function. Do this 1000 times and calculate the mean of each sample. Plot the sampling distributions of the means
using a histogram. Get the 2.5th and 97.5th percentiles of the
distribution. [Difficulty: **Beginner/Intermediate**] 
4. Use the `t.test()` function to calculate confidence intervals
of the mean on the first random sample `pois1` simulated from the `rpois()` function below. [Difficulty: **Intermediate**] 
```{r exRpoisChp3,eval=FALSE}
#HINT
set.seed(100)

#sample 30 values from poisson dist with lamda paramater =30
pois1=rpois(30,lambda=5)

```
5. Use the bootstrap confidence interval for the mean on `pois1`. [Difficulty: **Intermediate/Advanced**] 
6. Compare the theoretical confidence interval of the mean from the `t.test` and the bootstrap confidence interval. Are they similar? [Difficulty: **Intermediate/Advanced**]  
7. Try to re-create the following figure, which demonstrates the CLT concept.[Difficulty: **Advanced**] 
```{r,echo=FALSE,message=FALSE,warning=FALSE}
set.seed(101)
require(mosaic)
par(mfcol=c(4,3))
par(mar=c(5.1-2,4.1-1,4.1,2.1-2))
d=c(rnorm(1000,mean=10,sd=8))
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=10")),cex=0.6)
mtext("normal",cex=0.8,line=1)

bimod10=rowMeans(do(1000)*rnorm(5,mean=10,sd=8))
bimod30=rowMeans(do(1000)*rnorm(15,mean=10,sd=8))
bimod100=rowMeans(do(1000)*rnorm(50,mean=10,sd=8))
hist(bimod10,xlim=c(0,20),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=10",side=2,cex=0.8,line=2)
hist(bimod30,xlim=c(0,20),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=30",side=2,cex=0.8,line=2)
hist(bimod100,xlim=c(0,20),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("n=100",side=2,cex=0.8,line=2)

d=rexp(1000)
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=1")),cex=0.6)
mtext("exponential",cex=0.8,line=1)
mtext("Distributions of different populations",line=2)

exp10 =rowMeans(do(2000)*rexp(10))
exp30 =rowMeans(do(2000)*rexp(30))
exp100=rowMeans(do(2000)*rexp(100))
hist(exp10,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
mtext("Sampling distribution of sample means",line=2)
hist(exp30,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(exp100,xlim=c(0,2),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")

d=runif(1000)
hist(d,main="",
     col="black",border="white",breaks=20,xlab="",ylab=""
     )
abline(v=mean(d),col="red")
mtext(expression(paste(mu,"=0.5")),cex=0.6)

mtext("uniform",cex=0.8,line=1)
unif10 =rowMeans(do(1000)*runif(10))
unif30 =rowMeans(do(1000)*runif(30))
unif100=rowMeans(do(1000)*runif(100))
hist(unif10,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(unif30,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
hist(unif100,xlim=c(0,1),main="",xlab="",ylab="",breaks=20,col="gray",
     border="gray")
```

### How to test for differences in samples
1. Test the difference of means of the following simulated genes
using the randomization, `t-test()`, and `wilcox.test()` functions.
Plot the distributions using histograms and boxplots. [Difficulty: **Intermediate/Advanced**] 
```{r exRnorm1chp3,eval=FALSE}
set.seed(101)
gene1=rnorm(30,mean=4,sd=3)
gene2=rnorm(30,mean=3,sd=3)

```

2. Test the difference of the means of the following simulated genes
using the randomization, `t-test()` and `wilcox.test()` functions.
Plot the distributions using histograms and boxplots. [Difficulty: **Intermediate/Advanced**] 
```{r exRnorm2chp3,eval=FALSE}
set.seed(100)
gene1=rnorm(30,mean=4,sd=2)
gene2=rnorm(30,mean=2,sd=2)

```

3. We need an extra data set for this exercise. Read the gene expression data set as follows:
`gexpFile=system.file("extdata","geneExpMat.rds",package="compGenomRData") data=readRDS(gexpFile)`. The data has 100 differentially expressed genes. The first 3 columns are the test samples, and the last 3 are the control samples. Do 
a t-test for each gene (each row is a gene), and record the p-values.
Then, do a moderated t-test, as shown in section "Moderated t-tests" in this chapter, and record 
the p-values. Make a p-value histogram and compare two approaches in terms of the number of significant tests with the $0.05$ threshold.
On the p-values use FDR (BH), Bonferroni and q-value adjustment methods.
Calculate how many adjusted p-values are below 0.05 for each approach.
[Difficulty: **Intermediate/Advanced**] 

### Relationship between variables: Linear models and correlation

Below we are going to simulate X and Y values that are needed for the 
rest of the exercise.
```{r exLM1chp3,eval=FALSE}
# set random number seed, so that the random numbers from the text
# is the same when you run the code.
set.seed(32)

# get 50 X values between 1 and 100
x = runif(50,1,100)

# set b0,b1 and variance (sigma)
b0 = 10
b1 = 2
sigma = 20
# simulate error terms from normal distribution
eps = rnorm(50,0,sigma)
# get y values from the linear equation and addition of error terms
y = b0 + b1*x+ eps
```


1. Run the code then fit a line to predict Y based on X. [Difficulty:**Intermediate**] 
2. Plot the scatter plot and the fitted line. [Difficulty:**Intermediate**] 
3. Calculate correlation and R^2. [Difficulty:**Intermediate**] 
4. Run the `summary()` function and 
try to extract P-values for the model from the object
returned by `summary`. See `?summary.lm`. [Difficulty:**Intermediate/Advanced**] 
5. Plot the residuals vs. the fitted values plot, by calling the `plot()` 
function with `which=1` as the second argument. First argument
is the model returned by `lm()`. [Difficulty:**Advanced**] 


6. For the next exercises, read the data set histone modification data set. Use the following to get the path to the file:
```
hmodFile=system.file("extdata",
                    "HistoneModeVSgeneExp.rds",
                     package="compGenomRData")`
```
There are 3 columns in the dataset. These are measured levels of H3K4me3,
H3K27me3 and gene expression per gene. Once you read in the data, plot the scatter plot for H3K4me3 vs. expression. [Difficulty:**Beginner**] 

7. Plot the scatter plot for H3K27me3 vs. expression. [Difficulty:**Beginner**] 

8. Fit the model for prediction of expression data using: 1) Only H3K4me3 as explanatory variable, 2) Only H3K27me3 as explanatory variable, and 3) Using both H3K4me3 and H3K27me3 as explanatory variables. Inspect the `summary()` function output in each case, which terms are significant. [Difficulty:**Beginner/Intermediate**] 

10. Is using H3K4me3 and H3K27me3 better than the model with only H3K4me3? [Difficulty:**Intermediate**] 

11. Plot H3k4me3 vs. H3k27me3. Inspect the points that do not
follow a linear trend. Are they clustered at certain segments 
of the plot? Bonus: Is there any biological or technical interpretation
for those points? [Difficulty:**Intermediate/Advanced**] 

