---
title: 'Intro to R: Exercises'
author: "Altuna Akalin"
date: "22 Sep 2015"
output: 
  html_document:
          toc: true
          toc_depth: 1
          number_sections: true
---



# The setup

Set your working directory to the source file location in RStudio top menu:

'Session > Set Working Directory > To Source File Location'

if you have done it correctly you should see this script when you type dir()


```
> dir()
[1] "intro2R.exercises.html"
```

# Computations in R

##
Sum 2 and 3, use +

```{r,eval=FALSE,echo=FALSE}
2+3
```

## 
Take the square root of 36, use sqrt()

```{r,echo=FALSE,eval=FALSE}
sqrt(36)
```

## 
Take the log10 of 1000, use function `log10()`

```{r,echo=FALSE,eval=FALSE}
log10(1000)
```

## 
Take the log2 of 32, use function `log2()`


```{r,echo=FALSE,eval=FALSE}
log2(32)
```

## 
Assign the sum of 2,3 and 4 to variable x

```{r,echo=FALSE,eval=FALSE}
x =  2+3+4
x <- 2+3+4
```
##
Find the absolute value of  `5 - 145` using `abs()` function
```{r,echo=FALSE,eval=FALSE}
abs(5-145)
```

##
Calculate the square root of 625, divide it by 5 and assign it to variable `x`.
Ex: `y= log10(1000)/5`, the previous statement takes log10 of 1000, divides it 
by 5 and assigns the value to variable y

```{r,echo=FALSE,eval=FALSE}
x = sqrt(625)/5
 
```

##
Multiply the value you get from previous exercise with 10000, assign it variable x
Ex: `y=y*5`, multiplies y with 5 and assigns the value to y.

**KEY CONCEPT:** results of computations or arbitrary values can be stored in variables we can re-use those variables later on and over-write them with new values

```{r,echo=FALSE,eval=FALSE}
x2 = x*10000
```

# Data structures in R 


## 
Make a vector of 1,2,3,5 and 10 using `c()`, assign it to `vec` variable.
Ex: `vec1=c(1,3,4)` makes a vector out of 1,3,4.

```{r,echo=FALSE,eval=FALSE}
c(1:5,10)
vec1=c(1,2,3,4,5,10)
```

## 
Check the length of your vector with length(). 
Ex: `length(vec1)` should return 3

```{r,echo=FALSE,eval=FALSE}
length(vec1)
```

## 
Make a vector of all numbers between 2 and 15.
Ex: `vec=1:6` makes a vector of numbers between 1 and 6, assigns to vec variable

```{r,echo=FALSE,eval=FALSE}
vec=2:15 
```

##
Make a vector of  4s repeated 10 times using `rep()` function.
Ex: `rep(x=2,times=5)` makes a vector of 2s repeated 5 times

```{r,echo=FALSE,eval=FALSE}
rep(x=4,times=10)
rep(4,10)
```

## 
Make a logical vector with TRUE, FALSE values of length 4, use `c()`.
Ex: `c(TRUE,FALSE)`

```{r,echo=FALSE,eval=FALSE}
c(TRUE,FALSE,FALSE,TRUE,FALSE)
c(TRUE,TRUE,FALSE,TRUE,FALSE)
``` 

## 
Make a character vector of gene names PAX6,ZIC2,OCT4 and SOX2.
Ex: `avec=c("a","b","c")` a makes a character vector of a,b and c

```{r,echo=FALSE,eval=FALSE}
c("PAX6","ZIC2","OCT4","SOX2")
```

## 
Subset the vector using `[]` notation, get 5th and 6th elements. 
Ex: `vec1[1]` gets the first element. `vec1[c(1,3)]` gets 1st and 3rd elements

```{r,echo=FALSE,eval=FALSE}
vec1[c(5,6)]
```

## 
You can also subset any vector using a logical vector in `[]`.
Run the following:

```{r, eval=FALSE}
myvec=1:5
myvec[c(TRUE,TRUE,FALSE,FALSE,FALSE)] # the length of the logical vector should be equal to length(myvec) 
myvec[c(TRUE,FALSE,FALSE,FALSE,TRUE)]
``` 

## 
`==,>,<, >=, <=` operators create logical vectors. See the results of the following operations:

```{r,eval=FALSE}
myvec > 3
myvec == 4
myvec <= 2
myvec != 4
```

## 
Use `>` operator in `myvec[ ]` to get elements larger than 2 in `myvec` whic is described above
```{r,echo=FALSE,eval=FALSE}
myvec[ myvec > 2 ] 
myvec[ myvec == 2 ] 
myvec[ myvec != 2 ] 
```


##
make a 5x3 matrix (5 rows, 3 columns) using `matrix()`.
Ex: matrix(1:6,nrow=3,ncol=2) makes a 3x2 matrix using numbers between 1 and 6

```{r,echo=FALSE,eval=FALSE}

mat=matrix(1:15,nrow=5,ncol=3)

```

##
What happens when you use `byrow = TRUE` in your matrix() as an additional argument?
Ex: `mat=matrix(1:6,nrow=3,ncol=2,byrow = TRUE)`

```{r,echo=FALSE,eval=FALSE}
mat=matrix(1:15,nrow=5,ncol=3,byrow = TRUE)
```

## 
Extract first 3 columns and first 3 rows of your matrix using `[]` notation.

```{r,echo=FALSE,eval=FALSE}
mat[1:3,1:3]
```

## 
Extract last two rows of the matrix you created earlier.

Ex: `mat[2:3,]` or `mat[c(2,3),]` extracts 2nd and 3rd rows.

```{r,echo=FALSE,eval=FALSE}
mat[4:5,] 
mat[c(nrow(mat)-1,nrow(mat)),] 
tail(mat,n=1)
tail(mat,n=2)
```

##
Extract first two columns and run `class()` on the result.

```{r,echo=FALSE,eval=FALSE}
class(mat[,1:2])
```
##
Extract first column and run `class()` on the result, compare with the above exercise.

```{r,echo=FALSE,eval=FALSE}
class(mat[,1])
```

##
Make a data frame with 3 columns and 5 rows, make sure first column is sequence
of numbers 1:5, and second column is a character vector.

Ex: `df=data.frame(col1=1:3,col2=c("a","b","c"),col3=3:1) # 3x3 data frame`.

Remember you need to make 3x5 data frame

```{r,echo=FALSE,eval=FALSE}
df=data.frame(col1=1:5,col2=c("a","2","3","b","c"),col3=5:1)
```

##
Extract first two columns and first two rows. 

**HINT:** Same notation as matrices

```{r,echo=FALSE,eval=FALSE}
df[,1:2]

df[1:2,]
df[1:2,1:2]
```

## 
Extract last two rows of the data frame you made.

**HINT:** Same notation as matrices

```{r,echo=FALSE,eval=FALSE}
df[,4:5]
```
##
Extract last two columns using column names of the data frame you made.

```{r,echo=FALSE,eval=FALSE}
df[,c("col2","col3")]
```

## 
Extract second column using column names.
You can use `[]` or `$` as in lists, use both in two different answers.

```{r,echo=FALSE,eval=FALSE}
df$col2
df[,"col2"]
class(df["col1"])
class(df[,"col1"])
```
##
Extract rows where 1st column is larger than 3.
**HINT:** you can get a logical vector using `>` operator
,logical vectors can be used in `[]` when subsetting.

```{r,echo=FALSE,eval=FALSE}
df[df$col1 >3 , ]
```

##
Extract rows where 1st column is larger than or equal to 3.

```{r,echo=FALSE,eval=FALSE}
df[df$col1 >= 3 , ]
```

##
Convert data frame to the matrix. **HINT:** use `as.matrix()`.

Observe what happens to numeric values in the data frame.

```{r,echo=FALSE,eval=FALSE}
class(df[,c(1,3)])
as.matrix(df[,c(1,3)])
as.matrix(df)
```

## 
Make a list using `list()` function, your list should have 4 elements
the one below has 2.

Ex: `mylist= list(a=c(1,2,3),b=c("apple,"orange"))`

```{r,echo=FALSE,eval=FALSE}
mylist= list(a=c(1,2,3),
            b=c("apple","orange"),
            c=matrix(1:4,nrow=2),
            d="hello")
```

##
Select the 1st element of the list you made using `$` notation.

Ex: `mylist$a` selects first element named "a"

```{r,echo=FALSE,eval=FALSE}
mylist$a
```

## 
Select the 4th element of the list you made earlier using `$` notation.

```{r,echo=FALSE,eval=FALSE}
mylist$d
```

##
Select the 1st element of your list using `[ ]` notation. 

Ex: `mylist[1]` selects first element named "a", you get a list with one element.

Ex: `mylist["a"]` selects first element named "a", you get a list with one element.

```{r,echo=FALSE,eval=FALSE}
mylist[1] # -> still a list
mylist[[1]] # not a list

mylist["a"] 
mylist[["a"]] 
```

##
select the 4th element of your list using '[ ]' notation.

```{r,echo=FALSE,eval=FALSE}
mylist[4] 
mylist[[4]]
```

##
Make a factor using factor(), with 5 elements.
Ex: `fa=factor(c("a","a","b"))`

```{r,echo=FALSE,eval=FALSE}
fa=factor(c("a","a","b","c","c"))
```
##
Convert a character vector to factor using `as.factor()`.
First, make a character vector using `c()` then use `as.factor()`.

```{r,echo=FALSE,eval=FALSE}
my.vec=c("a","a","b","c","c")
fa=as.factor(my.vec)
fa
```
##
Convert the factor you made above to character using `as.character()`.

```{r,echo=FALSE,eval=FALSE}
fa
as.character(fa)
as.numeric()
as.factor()
```

# Reading in and writing data out in R

##
Read CpG island (CpGi) data from `../data/CpGi.table.hg18.txt`, this is a tab-separated file, store it in a variable called "cpgi".

```{r,echo=FALSE,eval=FALSE}
cpgi=read.table(file="../data/CpGi.table.hg18.txt",header=TRUE,sep="\t")
cpgi=read.table("../data/CpGi.table.hg18.txt",header=TRUE,sep="\t")
```
##
Use `head()` on CpGi to see first few rows.

```{r,echo=FALSE,eval=FALSE}
head(cpgi)
tail(cpgi)
```
##
Why doesn't the following work? see 'sep' argument at `help(read.table)`.

```{r}
cpgiSepComma=read.table("../data/CpGi.table.hg18.txt",header=TRUE,sep=",")
head(cpgiSepComma)
```

##
What happens when `stringsAsFactors=FALSE` ?
`cpgiHF=read.table("../data/CpGi.table.hg18.txt",header=FALSE,sep="\t",
                  stringsAsFactors=FALSE)`
                  
```{r,echo=FALSE,eval=FALSE}                  
head(cpgiHF)
head(cpgi)
class(cpgiHF$V2)
class(cpgiHF$V2)
```

##
Read only first 10 rows of the CpGi table.

```{r,echo=FALSE,eval=FALSE}
cpgi10row=read.table("../data/CpGi.table.hg18.txt",header=TRUE,sep="\t",nrow=10)
cpgi10row
```

##
Use `read.table("../data/CpGi.table.hg18.txt",...` with `header=FALSE`.

Do `head()` to see the results.

```{r,echo=FALSE,eval=FALSE}
df=read.table("../data/CpGi.table.hg18.txt",header=FALSE,sep="\t")
head(df)
```

##
Write CpG islands to a text file called "my.cpgi.file.txt". Write the file
to your home folder, you can use `file="~/my.cpgi.file.txt"` in linux. `~/` denotes
home folder.

```{r,echo=FALSE,eval=FALSE}
write.table(cpgi,file="~/my.cpgi.file.txt")
```

##
Same as above but this time make sure use `quote=FALSE`,`sep="\t"` and `row.names=FALSE` arguments. 
Save the file to "my.cpgi.file2.txt" and compare it with "my.cpgi.file.txt"

```{r,echo=FALSE,eval=FALSE}
write.table(cpgi,file="~/my.cpgi.file2.txt",quote=FALSE,sep="\t",row.names=FALSE)
```

##
Write out the first 10 rows of 'cpgi' data frame.
**HINT:** use subsetting for data frames we learned before.

```{r,echo=FALSE,eval=FALSE}
write.table(cpgi[1:10,],file="~/my.cpgi.fileNrow10.txt",quote=FALSE,sep="\t")
```

##
Write the first 3 columns of 'cpgi' data frame.

```{r,echo=FALSE,eval=FALSE}
dfnew=cpgi[,1:3]
write.table(dfnew,file="~/my.cpgi.fileCol3.txt",quote=FALSE,sep="\t")
```

##
Write CpG islands only on chr1.
**HINT:** use subsetting with `[]`, feed a logical vector using `==` operator.

```{r,echo=FALSE,eval=FALSE}
write.table(cpgi[cpgi$chrom == "chr1",],file="~/my.cpgi.fileChr1.txt",
            quote=FALSE,sep="\t",row.names=FALSE)
head(cpgi[cpgi$chrom == "chr1",])
```

##
Read two other data sets "../data/rn4.refseq.bed" and "../data/rn4.refseq2name.txt"
with `header=FALSE`, assign them to df1 and df2 respectively.

```{r,echo=FALSE,eval=FALSE}
df1=read.table("../data/rn4.refseq.bed",sep="\t",header=FALSE)
df2=read.table("../data/rn4.refseq2name.txt",sep="\t",header=FALSE)
```

##
Use `head()` to see what is inside of the the data frames above.
```{r,echo=FALSE,eval=FALSE}
head(df1)
head(df2)
```

##
Merge data sets using `merge()` and assign the results to variable named 'new.df', and use `head()` to see the results.

```{r,echo=FALSE,eval=FALSE}
new.df=merge(df1,df2,by.x="V4",by.y="V1")
head(new.df)
```


# Plotting in R


Please **run the following** for the rest of the exercises. 
```{r}
set.seed(1001)
x1=1:100+rnorm(100,mean=0,sd=15)
y1=1:100
```
##
Make a scatter plot using `x1` and `y1` vectors generated above.

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1)
```
##
Use `main` argument to give a title to `plot()` as in `plot(x,y,main="title")`

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,main="scatter plot")
```

##
Use `xlab` argument to set a label to x-axis.Use `ylab` argument to set a label to y-axis.
```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,main="scatter plot",xlab="x label")
```


##
See what `mtext(side=3,text="hi there")` does. **HINT:** mtext stands for margin text.

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,main="scatter plot",xlab="x label",ylab="y label")
mtext(side=3,text="hi there") 
```
##
See what `mtext(side=2,text="hi there")` does.check your plot after execution.

```{r,echo=FALSE,eval=FALSE}
mtext(side=2,text="hi there") 
mtext(side=4,text="hi there") 
```




##
You can use `paste()` as 'text' argument in mtext() try that, you need to re-plot.
your plot first. **HINT:** `mtext(side=3,text=paste(...))`
See what `paste()` is used for.

```{r}
paste("Text","here")
myText=paste("Text","here")
myText
```
Use *mtext()*  and *paste()* to put a margin text on the plot.

```{r,echo=FALSE,eval=FALSE}
mtext(side=3,text=paste("here","here"))
```
##
`cor()` calculates correlation between two vectors.
Pearson correlation is a measure of the linear correlation (dependence) 
between two variables X and Y.

```{r}
corxy=cor(x1,y1) # calculates pearson correlation
```

##
Try use `mtext()`,`cor()` and `paste()` to display correlation coefficient on
your scatterplot ?

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,main="scatter")
corxy=cor(x1,y1) 
#mtext(side=3,text=paste("Pearson Corr.",corxy))
mtext(side=3,text=paste("Pearson Corr.",round(corxy,3) ) )

plot(x1,y1)
mtext(side=3,text=paste("Pearson Corr.",round( cor(x1,y1)  ,3)  ) )
```

##
Change the colors of your plot using `col` argument.
Ex: `plot(x,y,col="red")`

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,col="red")
```

##
Use `pch=19` as an argument in your `plot()` command.

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,col="red",pch=19)
```

##
Use `pch=18` as an argument to your `plot()` command.

```{r,echo=FALSE,eval=FALSE}
plot(x1,y1,col="red",pch=18)
?points
```
##
Make histogram of x1 with `hist()` function.Histogram is a graphical representation of the data distribution.

```{r,echo=FALSE,eval=FALSE}
hist(x1)
```
##
You can change colors with 'col', add labels with 'xlab', 'ylab', and add a 'title' with 'main' arguments. Try all these in a histogram.
 
 ```{r,echo=FALSE,eval=FALSE}
hist(x1,main="title")
```

##
Make boxplot of y1 with `boxplot()`.

```{r,echo=FALSE,eval=FALSE}
boxplot(y1,main="title")
```

##
Make boxplots of `x1` and `y1` vectors in the same plot.

```{r,echo=FALSE,eval=FALSE}
boxplot(x1,y1,ylab="values",main="title")
```

##
In boxplot use horizontal = TRUE  argument

```{r,echo=FALSE,eval=FALSE}
boxplot(x1,y1,ylab="values",main="title",horizontal=TRUE)
```

##
make multiple plots with par(mfrow=c(2,1))
1. run par(mfrow=c(2,1))
2. make a boxplot 
3. make a histogram

```{r,echo=FALSE,eval=FALSE}
par( mfrow=c(1,2) )
hist(x1)
boxplot(y1)
```

##
Do the same as above but this time with par(mfrow=c(1,2))

```{r,echo=FALSE,eval=FALSE}
par(mfrow=c(2,2))
hist(x1)
boxplot(y1)
```

##
Save your plot using "Export" button


##
Save your plot by running :
```{r,eval=FALSE}
dev.copy(pdf,file="plot.file.pdf");dev.off()
```

##
Save your plot running :
```{r,eval=FALSE}
dev.copy(png,filename="plot.file.png");dev.off()
```

##
Another way to save the plot is the following

  1. Open a graphics device
  2. Create the plot
  3. Close the graphics device

```{r,echo=FALSE,eval=FALSE}
pdf("~/myplot1.pdf", width = 5, height = 5) # 1.
plot(x1, y1) # 2.
dev.off()   # 3.
```

##
**EXTRA: Making color density scatterplot**. You can make a scatter plot showing density
of points rather than points themselves. If you use points it looks like this:

```{r}

x2=1:1000+rnorm(1000,mean=0,sd=200)
y2=1:1000
plot(x2,y2,pch=19,col="blue")
```

If you use smoothScatter() function, you get the densities. 
```{r}
smoothScatter(x2,y2,colramp=colorRampPalette(c("white","blue", "green","yellow","red"))) 
```
 
Now, plot with `colramp=heat.colors` argument and then use a custom color scale using the following argument.
```
colramp = colorRampPalette(c("white","blue", "green","yellow","red")))
```

```{r,echo=FALSE,eval=FALSE}
smoothScatter(x2,y2,colramp = heat.colors )
smoothScatter(x2,y2,
               colramp = colorRampPalette(c("white","blue", "green","yellow","red")))
```

# Functions and control structures (for, if/else etc.)


##
Read CpG island data

```{r,eval=TRUE}
cpgi=read.table("../data/CpGi.table.hg18.txt",header=TRUE,sep="\t")
head(cpgi)
```
##
Check values at perGc column using a histogram.
'perGc' stands for GC percent => percentage of C+G nucleotides

```{r,echo=FALSE}
hist(cpgi$perGc) # most values are between 60 and 70
```

##
Make a boxplot for 'perGc' column

```{r,echo=FALSE}
boxplot(cpgi$perGc) 
```

##
Use if/else structure to decide if given GC percent high, low or medium.
If it is low, high, or medium. low < 60, high>75, medium is between 60 and 75
use greater or less than operators <  or  > . 
Fill in the values in the in code below, where it is written 'YOU_FILL_IN'

```{r,echo=TRUE,eval=FALSE}

GCper=65
 #result="low"# set initial value
  
  if(GCper < 60){ # check if GC value is lower than 60, assign "low" to result
    result="low"
    cat("low")
  }
  else if(GCper > 75){  # check if GC value is higher than 75, assign "high" to result
    result="high"
    cat("high")
  }else{ # if those two conditions fail then it must be "medium"
    result="medium"
  }

result

```

##
Write a function that takes a value of GC percent and decides
if it is low, high, or medium. low < 60, high>75, medium is between 60 and 75.
Fill in the values in the in code below, where it is written 'YOU_FILL_IN'



```
GCclass<-function(my.gc){
  
  YOU_FILL_IN
  
  return(result)
}
GCclass(10) # should return "low"
GCclass(90) # should return "high"
GCclass(65) # should return "medium"
```

```{r,echo=FALSE,eval=FALSE}
GCclass<-function(my.gc){
  
  result="low"# set initial value
  
  if(my.gc < 60){ # check if GC value is lower than 60, assign "low" to result
    result="low"
  }
  else if(my.gc > 75){  # check if GC value is higher than 75, assign "high" to result
    result="high"
  }else{ # if those two conditions fail then it must be "medium"
    result="medium"
  }
  return(result)
}
GCclass(10) # should return "low"
GCclass(90) # should return "high"
GCclass(65) # should return "medium"
```

##
Use a for loop to get GC percentage classes for gcValues below. Use the function
you wrote above.

```
gcValues=c(10,50,70,65,90)
for( i in YOU_FILL_IN){
  YOU_FILL_IN
}
```

```{r,echo=FALSE,eval=FALSE}
gcValues=c(10,50,70,65,90)
for( i in gcValues){
 
  print(GCclass(i) )
}
```

##
Use `lapply` to get to get GC percentage classes for gcValues.
Example: 

```{r,eval=FALSE}
vec=c(1,2,4,5)
power2=function(x){ return(x^2)  }
    lapply(vec,power2)
```

```{r,echo=FALSE,eval=FALSE}
s=lapply(gcValues,GCclass)
```
##
Use sapply to get values to get GC percentage classes for gcValues
```{r,echo=FALSE,eval=FALSE}
sapply(gcValues,GCclass)
```

##
Is there a way to decide on the GC percentage class of given vector of GCpercentages
without using if/else structure and loops ? if so, how can you do it?
**HINT:** subsetting using < and > operators

```{r,echo=FALSE,eval=FALSE}
result=rep("low",length(gcValues) )
result[gcValues > 75]="high"
result[gcValues < 75 & gcValues > 60 ] = "medium"
```
